### 2019-05-16 Article_AS.Rmd
firt commit

---
title: "On the eletrodynamic of moving bodies"
author: "André Dogninema Sora, Jean-Romain Roussel, Alexis Achim"
output: pdf_document
bibliography: biblio.bibtex
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

Conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties. Wood density is among the key traits as it is related to mechanical strength and dimensional stability. Variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties. In the present study, we aim at disentangling age, genetic and environmental influences on the pith-to-bark variation of wood density. 
We analyzed the wood density profiles of a 16-year-old genetic experimentation plantation from white spruce seedlings grown in nurseries and planted at two years of age. repeated in two contrasting environments (Asselin and Saint-Casimir) in the province of Quebec. A total of 2365 trees belonging to 93 full-sib families were sampled for this study. We developed a model to describe the interannual variation of the wood density from  pith-to-bark. Then, we developed a variability index from the model residues to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.. The genetic control of density, ring width,  model parameters and uniformity index was evaluated . This makes it possible to see the trait the level of heritability of each of these characters and thus to know if it is useful to select trees from orbservations made on these trees. ASReml3 package in R was used to estimate trait genetic parameters based on the residual maximum likelihood. Strong heritability has been observed in density, ring width and model parameters have heritability ranging from low to moderate, and the index of variability has almost no heritability. Future work can be done on older trees and also on one-time events to see if this changes the heritability of the variability index.\

# Introduction
Thousands of hectares of Canadian forest are harvested each year in addition to losses due to insects, diseases and fires. It is therefore essential to offset these losses if we maintain Canada's forests. Genetic improvement, which is a tool for reforestation, is becoming a crucial and increasingly necessary discipline for the development of the Canadian forest. Since trees are organisms that live for several decades, it is essential to take into account the current climate change context by using trees capable of adapting to current and future climatic conditions. It is in this context that the genetic improvement intervenes so that the reforestation is made from tree to adapt to the climate and having interesting characters for wood production. Improvement programs are in place for the most important commercial species, including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance. Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008]. In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@Nienstaedt1990]. All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly accross almost all forested regions which represents a wide range of soils types and a climatic conditions (Canadian Council of Forest Ministers 2009). \

During the first half of the 20th century, most studies were dedicated on increasing wood volume production [@Koubaa2000; REF] but with the increasing demand of high structural quality products the aspects of wood quality became as important as wood volume [@Franceschini2012]. Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood properties [@Corriveau1971; @Li1993; @Li1997]. On the other hand, it must be remembered that variations in certain wood properties such as ring width and density are governed by the environment in addition to genetics [@Dufour2010, @Rossi2012]. It is therefore essential to take climate conditions into account in breeding programs for wood quality. \

Wood quality has two components: resistance to mortality by vigor, for example, and quality for human use, which is often focused on density. Indeed a uniform density at the stem of the tree gives it a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm. With regard to the quality of wood related to its use, Several traits were considered for wood quality assessment and density is one of the most widely used indicators and arguably the most important one because it affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Macdonald2002; @Mckendry2002]. For exemple, a strong interannual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance. It is therefore important from both a biological point of view and use of the wood resource to have a wood with a uniform density. \

Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b, @Beaulieu2006] and this makes this trait attractive for genetic improvement. It should be noted that although density is predominantly under genetic control, it is partly controlled by the environment as growth is density dependent [@Pape1999] and growth is predominantly environmentally controlled.\

There are two types of radial variation in wood density: low frequency radial variations under internal control and high frequency interannual variations under control of annual events (environment) [@Nicault2010]. In white spruce, there is a typical radial pattern of variation in bark-to-bark wood density in spruce [@Moore2011], which is a low-frequency radial variation. First a decrease in density during the first years of growth and then an increase of from the beginning of the formation of mature wood. This typical pattern is denoted as pattern of type II according to according to [@panshin1980]. In addition to this radial variation, the density can vary from one species to another or from one tree to another, but it also varies within the same tree (interannual variation) [@Nicault2010]. Contrary to the radial variation, the interannual variation of the density is influenced mainly by events external to the tree such as the edaphic and climatic conditions [@Nicault2010].\

In addition, there is little known genetic control of wood density uniformity in white spruce. The assessment of this trait can facilitate the selection of trees for genetic improvement or reforestation. The radial trend in black spruce density has been modelled using models [@Xiang2014]. This model can be adapted to white spruce that has the same radial tendency as black spruce. \
In  this study, we tested  using a model similar to that of @Xiang2014 to subtract the effect of cambial age from wood density in order to assess the interannual variation related to environmental conditions. The specific objectives were: (1) present the development of a model which describes the general white spruce density annual ring number from the pith, (2) characterizing the genetic influence on wood density variation in white spruce. In this way, we will be able to identify tree families that are least sensitive to interannual variations in climate.

# Material and method

# Conclusion
This study assessed the variability of wood density in white spruce. Using the optimal function of the stats package to find the best parameters iteratively for each tree allowed to have an adjustment for the density profile of each of the trees in the study. Thus, the index of variability of the density was determined from the profile of each tree. The low heritability of this index puts a damper on its use for identifying potential tree family of interest for breeding programs from a low interannual variability perspective of wood density. As for the heritability of ring density and width, it is similar to what can be found in previous studies. The application of the method used in this study to older trees can be interesting in order to make a comparison and see if the heritability is the same and if this index can be used in this case to select families of trees for genetic improvement.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-05-16 Article_AS.Rmd
Suppresion espace
Commentaire
Reprise intro

@@
 title: "On the eletrodynamic of moving bodies"
-author: "André Dogninema Sora, Jean-Romain Roussel, Alexis Achim"
-output: pdf_document
+author: "André Dogninema Soro, Jean-Romain Roussel, Alexis Achim"
+output: 
+  pdf_document:
+classoption: twocolumn
 bibliography: biblio.bibtex
@@
 We developed a model to describe the interannual variation of the wood density from  pith-to-bark.
-Then, we developed a variability index from the model residues to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions..
-The genetic control of density, ring width,  model parameters and uniformity index was evaluated .
+Then, we developed a variability index from the model residues to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
+The genetic control of density, ring width,  model parameters and uniformity index was evaluated.
 This makes it possible to see the trait the level of heritability of each of these characters and thus to know if it is useful to select trees from orbservations made on these trees.
@@
 
+\clearpage
+
 # Introduction
-Thousands of hectares of Canadian forest are harvested each year in addition to losses due to insects, diseases and fires.
-It is therefore essential to offset these losses if we maintain Canada's forests.
-Genetic improvement, which is a tool for reforestation, is becoming a crucial and increasingly necessary discipline for the development of the Canadian forest.
-Since trees are organisms that live for several decades, it is essential to take into account the current climate change context by using trees capable of adapting to current and future climatic conditions.
-It is in this context that the genetic improvement intervenes so that the reforestation is made from tree to adapt to the climate and having interesting characters for wood production.
-Improvement programs are in place for the most important commercial species, including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance.
+
+Thousands of hectares of canadian forest are harvested each year and an additionnal loss is due to insects, diseases and fires [@X, @Y].
+To maintain ressource constant, it is therefore essential to compensate the sampling and the looses by reforesting.
+The reforestation, when under human control, could ideally be made by using better trees than the former ones, i.e.
+trees that growth faster, that are more resistant to perturbation and that are of better stuctural quality.
+To achieve this task, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@X, @Y].
+\
+
+Improvement programs by genetic selection are in place for the most important commercial species [@X,@Y] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@X,@Y].
 Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008].
@@
 
-During the first half of the 20th century, most studies were dedicated on increasing wood volume production [@Koubaa2000; REF] but with the increasing demand of high structural quality products the aspects of wood quality became as important as wood volume [@Franceschini2012].
-Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood properties [@Corriveau1971; @Li1993; @Li1997].
-On the other hand, it must be remembered that variations in certain wood properties such as ring width and density are governed by the environment in addition to genetics [@Dufour2010, @Rossi2012].
-It is therefore essential to take climate conditions into account in breeding programs for wood quality.
+During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Koubaa2000; @X,@Y] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Li1997].
+Wood quality has mainly two components: biologic and economic.
+The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
+Several traits were considered for wood quality assessment and the density is one of the most widely used indicators and arguably the most important one [@X, @Y] because this trait is important for both components of quality.
 \
 
-Wood quality has two components: resistance to mortality by vigor, for example, and quality for human use, which is often focused on density.
-Indeed a uniform density at the stem of the tree gives it a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
-With regard to the quality of wood related to its use, Several traits were considered for wood quality assessment and density is one of the most widely used indicators and arguably the most important one because it affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Macdonald2002; @Mckendry2002].
+With regard to the biological component, in has been demonstrated [@X, @Y] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
+For example a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
+With regard to the economical context, uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Macdonald2002; @Mckendry2002].
 For exemple, a strong interannual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance.
-It is therefore important from both a biological point of view and use of the wood resource to have a wood with a uniform density.
+It is therefore important from both a biological point of view and a end-use point of view to have a wood with a uniform density within the stem.
+\
+
+From these studies we understand that there might be a strong gain at selecting trees that are prone to produce wood of uniform density to get a resistant and higly valuable forest in the next decades.
+To reach this goal is is important to understant where do the variations of density come from.
+In white spruce, there is a typical low frequency radial pattern of variation from pith to bark [@Moore2011].
+First a decrease of density during the first years of growth and then an increase from the beginning of the formation of mature wood [@panshin1980].
+In addition this radial variations there are hight frequency interannual variation [@Nicault2010].
+Contrary to the radial variation that is under internal control, the interannual variation of the density is influenced mainly by events events to the tree such as the edaphic and climatic conditions [@Nicault2010] that happen at high frequency.
 \
@@
 Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b, @Beaulieu2006] and this makes this trait attractive for genetic improvement.
-It should be noted that although density is predominantly under genetic control, it is partly controlled by the environment as growth is density dependent [@Pape1999] and growth is predominantly environmentally controlled.\
 
-There are two types of radial variation in wood density: low frequency radial variations under internal control and high frequency interannual variations under control of annual events (environment) [@Nicault2010].
-In white spruce, there is a typical radial pattern of variation in bark-to-bark wood density in spruce [@Moore2011], which is a low-frequency radial variation.
-First a decrease in density during the first years of growth and then an increase of from the beginning of the formation of mature wood.
-This typical pattern is denoted as pattern of type II according to according to [@panshin1980].
-In addition to this radial variation, the density can vary from one species to another or from one tree to another, but it also varies within the same tree (interannual variation) [@Nicault2010].
-Contrary to the radial variation, the interannual variation of the density is influenced mainly by events external to the tree such as the edaphic and climatic conditions [@Nicault2010].\
+<!-- Ca dit tout l'inverse de ce que veux dire... je sais quoi faire de ca. -->
+<!-- It should be noted that although density is predominantly under genetic control, it is partly controlled by the environment as growth is density dependent [@Pape1999] and growth is predominantly environmentally controlled.
+\ -->
 
-In addition, there is little known genetic control of wood density uniformity in white spruce.
-The assessment of this trait can facilitate the selection of trees for genetic improvement or reforestation.
-The radial trend in black spruce density has been modelled using models [@Xiang2014].
-This model can be adapted to white spruce that has the same radial tendency as black spruce.
+However, the timeframe of such improved forest management being long it is essential to take into account the current climate change context [@X,@Y].
+With the increasing temperature, the trees are expected to growth faster and thus being of lower quality [@X,@Y].
+In addition to selecting trees that are prone to produce wood of uniform density we must also ensure that these properties will persist in time with climatic change.
+Thus the selection must also considered the stability of the properties in time or, more accuralty, with climate change.
 \
-In  this study, we tested  using a model similar to that of @Xiang2014 to subtract the effect of cambial age from wood density in order to assess the interannual variation related to environmental conditions.
+
+In this study we quantified and compared heteregoneity of wood density from trees of known parents in order to figure out if some individus produce more uniform wood than others.
+Knowing the parents of each tree we tried to characterize the genetic contribution to the homogeneity.
+By comparing two growth sites we tried to characterize the environmental contribution to homogeity.
+We ultimalty aim to select and clone individus of interest that are both insensitive to climate change and genetically prone to make homogenous wood.
+
+<!-- objectif pas clair: toujours la même question: pourquoi ? -->
+<!-- In  this study, we tested using a model similar to that of @Xiang2014 to subtract the effect of cambial age from wood density in order to assess the interannual variation related to environmental conditions.
 The specific objectives were: (1) present the development of a model which describes the general white spruce density annual ring number from the pith, (2) characterizing the genetic influence on wood density variation in white spruce.
 In this way, we will be able to identify tree families that are least sensitive to interannual variations in climate.
+-->
+
+\clearpage
 
 # Material and method
+
+\clearpage
 
@@
 Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+
+\clearpage
+
+# References

### 2019-05-16 Article_AS.Rmd
Materiel et methodes

--- 2019-05-16 Article_AS.Rmd
@@
 # Abstract
+
+<!-- JR: Commentaire -->
 
@@
 # Introduction
+
 Thousands of hectares of Canadian forest are harvested each year in addition to losses due to insects, diseases and fires.
@@
 
+## Sampling
+The dataset used in this study came from genetic selection for a series of tests according to provenances [@Beaulieu1996].
+The improvement program for this region was led by Quebec’s Ministry of Forests, Wildlife and Parcs, Canada.
+Experimental plantings were initiated from white spruce seedlings grown in nurseries and planted at two years of age at a regular spacing between trees of 2 m in all directions.
+This series of tests was established in 1999 by the Canadian Forest Service and is one of the three essential parts of the white spruce breeding program in Quebec.
+Two sites were selected for sampling: one near Saint-Casimir, Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather).
+Between nine and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees (Table 1).
+Trees were 15 years old at the time of sampling.
+Trees were assigned to row-plots of five trees/plot (2 m × 2 m spacing).
+Core samples were extracted at breast height (1.3 m above ground level) from the southern side of the trees to minimize the variation due to sample orientation.
+Tree height and diameter at breast height (DBH, 1.3 m above ground) were recorded in 2015 at the age of 15 years, and a wood increment core was extracted from the south facing side of each tree.
+Cores were stored in a freezer, conditioned to 7% moisture and cut to 1.68 mm thickness prior to X-ray densitometry analyses (Quintek Measurement Systems, TN).
+Wood density was calculated as a ring area weighted mean from recorded pith to bark wood density profiles.
+
+Study sites (image)
+
+Growth ring density and ring width measurements were taken from pith-to-bark, using an X-ray densitometer (Quintek Measurement Systems, TN) at a resolution of 0.0254 mm, and the data were reported as relative wood density on a dry weight basis.
+To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983].
+
+## Model Development
+A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce. 
+Considering the hierarchical structure of the data, with annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
+This type of model also considers the autocorrelation between the observations [@Lindstrom1990].
+Thus, the estimates of the parameters can vary randomly around the average values of each level of the data structure (site, block, tree, core) [@Auty2013].
+As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores.
+A nonlinear equation was fit for each core to describe ring density profiles, producing different parameter estimates for each core.
+The radial pattern of wood density in white spruce can be described as follows:
+
+```
+RD = b0*e ( b1/CA ) + CA/b2 + e
+
+```
+where RD is the mean ring density (kg∙m-3), CA is cambial age (years), b0, b1 and b2 are empirically determined parameters, and e is the residual error term.
+Parameter b0 denotes the intercept when CA approaches a theoretical value of 0.
+We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
+We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
+The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R2) (Parresol 1999).
+The graphic description in figure 1 of the three parameters of our model makes it possible to
+
 # Conclusion

### 2019-05-16 Article_AS.Rmd

--- 2019-05-16 Article_AS.Rmd
@@
 
+## Sampling
+The dataset used in this study came from genetic selection for a series of tests according to provenances [@Beaulieu1996].
+The improvement program for this region was led by Quebec’s Ministry of Forests, Wildlife and Parcs, Canada.
+Experimental plantings were initiated from white spruce seedlings grown in nurseries and planted at two years of age at a regular spacing between trees of 2 m in all directions.
+This series of tests was established in 1999 by the Canadian Forest Service and is one of the three essential parts of the white spruce breeding program in Quebec.
+Two sites were selected for sampling: one near Saint-Casimir, Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather).
+Between nine and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees (Table 1).
+Trees were 15 years old at the time of sampling.
+Trees were assigned to row-plots of five trees/plot (2 m × 2 m spacing).
+Core samples were extracted at breast height (1.3 m above ground level) from the southern side of the trees to minimize the variation due to sample orientation.
+Tree height and diameter at breast height (DBH, 1.3 m above ground) were recorded in 2015 at the age of 15 years, and a wood increment core was extracted from the south facing side of each tree.
+Cores were stored in a freezer, conditioned to 7% moisture and cut to 1.68 mm thickness prior to X-ray densitometry analyses (Quintek Measurement Systems, TN).
+Wood density was calculated as a ring area weighted mean from recorded pith to bark wood density profiles.
+
+Study sites (image)
+
+Growth ring density and ring width measurements were taken from pith-to-bark, using an X-ray densitometer (Quintek Measurement Systems, TN) at a resolution of 0.0254 mm, and the data were reported as relative wood density on a dry weight basis.
+To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983].
+
+## Model Development
+A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce. 
+Considering the hierarchical structure of the data, with annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
+This type of model also considers the autocorrelation between the observations [@Lindstrom1990].
+Thus, the estimates of the parameters can vary randomly around the average values of each level of the data structure (site, block, tree, core) [@Auty2013].
+As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores.
+A nonlinear equation was fit for each core to describe ring density profiles, producing different parameter estimates for each core.
+The radial pattern of wood density in white spruce can be described as follows:
+
+```
+RD = b0*e ( b1/CA ) + CA/b2 + e
+
+```
+where RD is the mean ring density (kg∙m-3), CA is cambial age (years), b0, b1 and b2 are empirically determined parameters, and e is the residual error term.
+Parameter b0 denotes the intercept when CA approaches a theoretical value of 0.
+We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
+We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
+The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R2) (Parresol 1999).
+The graphic description in figure 1 of the three parameters of our model makes it possible to
+
 \clearpage

### 2019-05-16 Article_AS.Rmd
materiel et methode

@@
 ```
-where RD is the mean ring density (kg∙m-3), CA is cambial age (years), b0, b1 and b2 are empirically determined parameters, and e is the residual error term.
+where RD is the mean ring density (kgm-3), CA is cambial age (years), b0, b1 and b2 are empirically determined parameters, and e is the residual error term.
 Parameter b0 denotes the intercept when CA approaches a theoretical value of 0.

### 2019-05-16 Article_AS.Rmd
Ajout Mat et met

@@
 The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R2) (Parresol 1999).
-The graphic description in figure 1 of the three parameters of our model makes it possible to
+The graphic description in figure 1 of the three parameters of our model makes it possible to understand how each parameter influences our model (Figure 1).
+
+![paramtres](/home/andre/Travail/Thèse/Article_AS/graph_table/parameters.png)
+
+Figure 1 : Graphic of description of different parameters b0, b1 and b2 of the model.
+Increasing the value of the parameters b0 and b1 results in an increase in the density at the level of all the rings.
+The increase in the parameter b2 causes a sharp rise in density after the decrease in the first rings.
+
+It can be observed in figure 1 that the parameters b0 and b1 influence the whole of the curve.
+An increase of one of these parameters leads to that of the density on all rings.
+Unlike parameter b0 and b1, parameter b2 seems to be influenced by some of the density.
+A low value of this parameter causes a sharp increase of the density starting from the third ring and when this parameter increases a decrease of the density is observed at this level until the curve is similar to the cases of the parameters b0 and b1.
+Beyond the value 1, the parameter b2 does not influence the variation of the density.\
+
+## Determination of the wood density uniformity index
+The uniformity index is an index to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
+The residues of our model were used to obtain the wood density uniformity index, which was determined using an equation.
+We subtracted the density residues values of year n-1 from year n.
+We were interested more specifically in the difference between the average density residues of annual rings of two consecutive years, no matter whether the difference was positive or negative.
+The absolute value was used for each difference to get rid of the sign.
+The index was estimated from the following formula:
+
+```
+H = ∑ abs ( Dn − Dn − 1 )
+
+```
+Where H is the uniformity index, Dn  is the wood density residues of year N and  Dn − 1 is the tree density residues of year N-1.
+
+## Spearman’s rank correlation coefficient (Spearman’s rho) 
+To describe the link between two variables by means of a monotonic function, we used Spearman’s rank correlation coefficient.
+We calculated this coefficient for the index of uniformity and the different parameters of the model.
+The elements for which we calculated Spearman’s rho are: the uniformity index for each of the tree families between the two study sites, the model residues for each of the families and the three parameters to estimate from the model.
+The cor.test function in the R (Team Core 2018 R) statistical programming environment was used to calculate Spearman’s rho.
+We used a 95 percent confidence interval.
+
+## Quantitative genetic analyses
+Variance and covariance for genetics were estimated using mixed models (Lenz et al., 2013) of the R software. The model used is as follows:
+
+```
+yijkl = μ + si + b(i)j + fk + bf (i)jk + eijkl
+
+```
+
+where yijkl  is the observation on the ijklth tree, μ is the overall mean, si is the fixed site effect, b(i)j is the fixed effect of the j th bloc in site i, fk is the random effect of the k th family, bf(i)j is the random interaction of family and bloc, and e ijkl is the residual error.
+In a full-sib design with random mating, members of the same family are presumed to share half of their genes (White et al., 2007). Therefore, individual heritability was obtained as follows:
+
+```
+h2  = 2*Vf / VT   
+```
+
+Where h2 is individual heritability, Vf is family variance, VT is total variance.
+ASReml3 package (David Butler, 2018) in the R statistical programming environment (R Core Team 2018) was used to estimate trait genetic parameters based on the residual maximum likelihood.
 

### 2019-05-21 Article_AS.Rmd
Ajout de biblio et modif d'intro

@@
 output: 
-  pdf_document:
+ html_document:
 classoption: twocolumn
@@
 
-Thousands of hectares of canadian forest are harvested each year and an additionnal loss is due to insects, diseases and fires [@X, @Y].
+Canadian forest is an important part of the country's history, culture, natural environment and economy.
+But thousands of hectares of canadian forest are harvested each year and an additionnal loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
 To maintain ressource constant, it is therefore essential to compensate the sampling and the looses by reforesting.
@@
 trees that growth faster, that are more resistant to perturbation and that are of better stuctural quality.
-To achieve this task, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@X, @Y].
+To achieve this task, we must take into account the cambial activity that is governed by environmental conditions and genetics [@Dufour2010; @Rossi2012].
+Therefore, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@Bonfils2004].
 \
 
-Improvement programs by genetic selection are in place for the most important commercial species [@X,@Y] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@X,@Y].
+Improvement programs by genetic selection are in place for the most important commercial species [@DRF2000; @Mullin2011] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015].
 Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008].
@@
 
-During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Koubaa2000; @X,@Y] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
-Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Li1997].
+During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
 Wood quality has mainly two components: biologic and economic.
 The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
-Several traits were considered for wood quality assessment and the density is one of the most widely used indicators and arguably the most important one [@X, @Y] because this trait is important for both components of quality.
+Several traits were considered for wood quality assessment and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006] because this trait is important for both components of quality.
 \
 
-With regard to the biological component, in has been demonstrated [@X, @Y] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
+With regard to the biological component, in has been demonstrated [@Hietz2013, @osazuwa2014] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
 For example a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
-With regard to the economical context, uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Macdonald2002; @Mckendry2002].
+With regard to the economical context, uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
 For exemple, a strong interannual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance.
@@
 
-<!-- Ca dit tout l'inverse de ce que veux dire... je sais quoi faire de ca. -->
-<!-- It should be noted that although density is predominantly under genetic control, it is partly controlled by the environment as growth is density dependent [@Pape1999] and growth is predominantly environmentally controlled.
-\ -->
-
-However, the timeframe of such improved forest management being long it is essential to take into account the current climate change context [@X,@Y].
-With the increasing temperature, the trees are expected to growth faster and thus being of lower quality [@X,@Y].
+However, the timeframe of such improved forest management being long [@Zobel1972] it is essential to take into account the current climate change context [@Gray2016].
+With the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005].
 In addition to selecting trees that are prone to produce wood of uniform density we must also ensure that these properties will persist in time with climatic change.
@@
 By comparing two growth sites we tried to characterize the environmental contribution to homogeity.
-We ultimalty aim to select and clone individus of interest that are both insensitive to climate change and genetically prone to make homogenous wood.
-
-<!-- objectif pas clair: toujours la même question: pourquoi ? -->
-<!-- In  this study, we tested using a model similar to that of @Xiang2014 to subtract the effect of cambial age from wood density in order to assess the interannual variation related to environmental conditions.
-The specific objectives were: (1) present the development of a model which describes the general white spruce density annual ring number from the pith, (2) characterizing the genetic influence on wood density variation in white spruce.
-In this way, we will be able to identify tree families that are least sensitive to interannual variations in climate.
--->
+We ultimalty aim to select and reproduce trees families of interest that are both  low-sensitivity to climate change and genetically prone to make homogenous wood.
 
@@
 
-![paramtres](/home/andre/Travail/Thèse/Article_AS/graph_table/parameters.png)
+ Graphe parametre
 

### 2019-05-21 Article_AS.Rmd
Math

@@
 output: 
- html_document:
+ pdf_document:
 classoption: twocolumn
@@
 
-```
-RD = b0*e ( b1/CA ) + CA/b2 + e
-
-```
-where RD is the mean ring density (kgm-3), CA is cambial age (years), b0, b1 and b2 are empirically determined parameters, and e is the residual error term.
-Parameter b0 denotes the intercept when CA approaches a theoretical value of 0.
+$$RD = b_0 \times \exp\left(\frac{b_1}{CA}\right) + \frac{CA}{b_2} + e $$
+where RD is the mean ring density (kgm-3), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and e is the residual error term.
+Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.
 We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
@@
 
-Figure 1 : Graphic of description of different parameters b0, b1 and b2 of the model.
-Increasing the value of the parameters b0 and b1 results in an increase in the density at the level of all the rings.
-The increase in the parameter b2 causes a sharp rise in density after the decrease in the first rings.
+Figure 1 : Graphic of description of different parameters $b_0$, $b_1$ and $b_2$ of the model.
+Increasing the value of the parameters $b_0$ and $b_1$ results in an increase in the density at the level of all the rings.
+The increase in the parameter $b_2$ causes a sharp rise in density after the decrease in the first rings.
 
-It can be observed in figure 1 that the parameters b0 and b1 influence the whole of the curve.
+It can be observed in figure 1 that the parameters $b_0$ and $b_1$ influence the whole of the curve.
 An increase of one of these parameters leads to that of the density on all rings.
-Unlike parameter b0 and b1, parameter b2 seems to be influenced by some of the density.
-A low value of this parameter causes a sharp increase of the density starting from the third ring and when this parameter increases a decrease of the density is observed at this level until the curve is similar to the cases of the parameters b0 and b1.
-Beyond the value 1, the parameter b2 does not influence the variation of the density.\
+Unlike parameter $b_0$ and $b_1$, parameter $b_2$ seems to be influenced by some of the density.
+A low value of this parameter causes a sharp increase of the density starting from the third ring and when this parameter increases a decrease of the density is observed at this level until the curve is similar to the cases of the parameters $b_0$ and $b_1$.
+Beyond the value 1, the parameter $b_2$ does not influence the variation of the density.\
 
@@
 The residues of our model were used to obtain the wood density uniformity index, which was determined using an equation.
-We subtracted the density residues values of year n-1 from year n.
+We subtracted the density residues values of year $n-1$ from year $n$.
 We were interested more specifically in the difference between the average density residues of annual rings of two consecutive years, no matter whether the difference was positive or negative.
@@
 
-```
-H = ∑ abs ( Dn − Dn − 1 )
+$$H = \sum | D_n - D_{n - 1} |$$
 
-```
-Where H is the uniformity index, Dn  is the wood density residues of year N and  Dn − 1 is the tree density residues of year N-1.
+Where H is the uniformity index, $D_n$ is the wood density residues of year $N$ and  $D_n - 1$ is the tree density residues of year $N-1$.
 
@@
 
-```
-yijkl = μ + si + b(i)j + fk + bf (i)jk + eijkl
+$$ y_{ijkl} = \mu + s_i + b_{(i)j} + f_k + bf_{(i)jk} + e_{ijkl}$$
 
-```
-
-where yijkl  is the observation on the ijklth tree, μ is the overall mean, si is the fixed site effect, b(i)j is the fixed effect of the j th bloc in site i, fk is the random effect of the k th family, bf(i)j is the random interaction of family and bloc, and e ijkl is the residual error.
+where yijkl  is the observation on the ijklth tree, $\mu$ is the overall mean, si is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j$ th bloc in site $i$, $f_k$ is the random effect of the k th family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl} is the residual error.
 In a full-sib design with random mating, members of the same family are presumed to share half of their genes (White et al., 2007). Therefore, individual heritability was obtained as follows:
 
-```
-h2  = 2*Vf / VT   
-```
+ $$h_2= 2 \frac{V_f}{V_T}$$
 
-Where h2 is individual heritability, Vf is family variance, VT is total variance.
+Where $h_2$ is individual heritability, $V_f$ is family variance, $V_T$ is total variance.
 ASReml3 package (David Butler, 2018) in the R statistical programming environment (R Core Team 2018) was used to estimate trait genetic parameters based on the residual maximum likelihood.

### 2019-05-23 Article_AS.Rmd
Figure
Reulstats et discussion

@@
 
-Study sites (image)
+![legend of this figure](graph_table/parameters.png)
 
@@
 
-where yijkl  is the observation on the ijklth tree, $\mu$ is the overall mean, si is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j$ th bloc in site $i$, $f_k$ is the random effect of the k th family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl} is the residual error.
+where yijkl  is the observation on the ijklth tree, $\mu$ is the overall mean, si is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j$ th bloc in site $i$, $f_k$ is the random effect of the k th family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
 In a full-sib design with random mating, members of the same family are presumed to share half of their genes (White et al., 2007). Therefore, individual heritability was obtained as follows:

### 2019-05-23 Article_AS.Rmd
Resultats et discussion2

--- 2019-05-21 Article_AS.Rmd
@@
 author: "André Dogninema Soro, Jean-Romain Roussel, Alexis Achim"
-output: 
- pdf_document:
-classoption: twocolumn
-bibliography: biblio.bibtex
 editor_options:
   chunk_output_type: inline
+output: pdf_document
+bibliography: biblio.bibtex
 ---
@@
 
-\clearpage
-
 # Material and method
@@
 Two sites were selected for sampling: one near Saint-Casimir, Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather).
-Between nine and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees (Table 1).
+Between nine and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees (fig 1).
 Trees were 15 years old at the time of sampling.
@@
 
-Study sites (image)
+![Study sites](graph_table/Study.sites.png){width=320px}
 
@@
 $$RD = b_0 \times \exp\left(\frac{b_1}{CA}\right) + \frac{CA}{b_2} + e $$
+
 where RD is the mean ring density (kgm-3), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and e is the residual error term.
@@
 We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
-The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R2) (Parresol 1999).
-The graphic description in figure 1 of the three parameters of our model makes it possible to understand how each parameter influences our model (Figure 1).
+The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R2) [@Pape1999].
+The graphic description in figure 1 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).
 
- Graphe parametre
-
-Figure 1 : Graphic of description of different parameters $b_0$, $b_1$ and $b_2$ of the model.
+![Graphic of description of different parameters $b_0$, $b_1$ and $b_2$ of the model.
 Increasing the value of the parameters $b_0$ and $b_1$ results in an increase in the density at the level of all the rings.
-The increase in the parameter $b_2$ causes a sharp rise in density after the decrease in the first rings.
+The increase in the parameter $b_2$ causes a sharp rise in density after the decrease in the first rings.](graph_table/parameters.pdf)
 
@@
 The elements for which we calculated Spearman’s rho are: the uniformity index for each of the tree families between the two study sites, the model residues for each of the families and the three parameters to estimate from the model.
-The cor.test function in the R (Team Core 2018 R) statistical programming environment was used to calculate Spearman’s rho.
+The cor.test function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho.
 We used a 95 percent confidence interval.
@@
 ## Quantitative genetic analyses
-Variance and covariance for genetics were estimated using mixed models (Lenz et al., 2013) of the R software. The model used is as follows:
+Variance and covariance for genetics were estimated using mixed models [@Lenz2013] of the R software. The model used is as follows:
 
@@
 
-where yijkl  is the observation on the ijklth tree, $\mu$ is the overall mean, si is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j$ th bloc in site $i$, $f_k$ is the random effect of the k th family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl} is the residual error.
-In a full-sib design with random mating, members of the same family are presumed to share half of their genes (White et al., 2007). Therefore, individual heritability was obtained as follows:
+where $yijkl$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
+In a full-sib design with random mating, members of the same family are presumed to share half of their genes [@white2007]. Therefore, individual heritability was obtained as follows:
 
@@
 Where $h_2$ is individual heritability, $V_f$ is family variance, $V_T$ is total variance.
-ASReml3 package (David Butler, 2018) in the R statistical programming environment (R Core Team 2018) was used to estimate trait genetic parameters based on the residual maximum likelihood.
+ASReml3 package [@Butler2018] in the R statistical programming environment [@R2018] was used to estimate trait genetic parameters based on the residual maximum likelihood.
 
-\clearpage
+#3 Results
+
+The model we have developed has made it possible to optimize the density of the wood for each study site.
+In table 1, we observe that the average value of the parameters differs between Asselin and Saint-Casimir.
+for the parameters b0 and b1, the average is higher in Saint-Casimir and for parameter b2 it is higher in Asselin.\
+
+|Site                     |Parameter |    Min|   Mean|    Max|
+|:------------------------|:---------|------:|------:|------:|
+| Asselin (Témiscouata)   |$b_0$     | 257.26| 302.78| 369.78|
+|                         |$b_1$     |  -0.93|   0.97|   5.36|
+|                         |$b_2$     | -81.83|  17.02|  61.34|
+| Saint-Casimir (Portneuf)|$b_0$     | 264.88| 304.40| 353.90|
+|                         |$b_1$     |  -1.34|   1.09|   7.04|
+|                         |$b_2$     | -50.87|  14.87|  61.34|
+Table: Summary table of variability index
+
+When considering all the cores of trees from the pith-to-bark, there is initially a rapid decrease in density in the first five years.
+It goes from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to kg/m^3^ for the Asselin site.
+Then, this decrease was followed by a slight increase until the twelfth year.\
+
+![The ring density versus the number of annual rings of the samples for the Asselin site.The bold lines represent the model prediction](graph_table/Density_asselin.png){width=330px; height=330px}
+
+![The ring density versus the number of annual rings of the samples for the Saint-Casimir site.The bold lines represent the model prediction](graph_table/Density_casimir.png){width=330px; height=330px}
+
+Model predictions are similar to data trends across sites (fig3&4).
+A decrease in density is observed in the first years of growth and then an increase in density from a certain cambial age.
+Whether it is the data or the predictions of the model, we observe a difference between the Asselin (fig 3) and Saint-Casimir (fig 4) sites.
+The increase in density after the decrease of the first years of growth is greater in Saint-Casimir than in Asselin.
+In spite of the interannual variations, the trees have the same general tendency and this one seems well described by the model which adapts to each of the sites.\
+
+<!--
+```{r aa, echo=FALSE}
+data <- readr::read_csv("data/Tableau_resume_indice d'uniformite.csv")
+knitr::kable(data)
+```
+-->
+
+|Site                     |  Min|  Mean|    Max|
+|:------------------------|----:|-----:|------:|
+|Asselin (Témiscouata)    |  0.0| 417.5| 1155.8|
+|Saint-Casimir (Portneuf) | 54.9| 339.8|  746.0|
+|Combined                 |  0.0| 375.9| 1155.8|
+Table: Summary table of variability index
+
+|Rank correlation  |  Rho| P-value|
+|:-----------------|----:|-------:|
+|Variability index | 0.08|    0.43|
+|$b_0$             | 0.70|    0.00|
+|$b_1$             | 0.80|    0.00|
+|$b_2$             | 0.60|    0.00|
+Table: Sperman’s Rang correlation
+
+The results of the rank correlation coefficient are summarized in Table 3.
+These results show an absence of correlation for the variability index (p-value = 0.43) and a strong correlation for all the parameters of the model (p-value = 2.2 10^-16^).
+The value of Rho is very similar between the three parameters.
+
+|Trait            |Combined sites |Asselin |Saint-Casimir |
+|:----------------|:--------------|:-------|:-------------|
+|Wood density     |0.55           | 0.58   |0.70          |
+|Width            |0.23           |0.32    |0.43          |
+|$b_0$            |0.20           |0.20    |0.3           |
+|$b_1$            |0.06           |0.10    |0.06          |
+|$b_2$            |0.04           |0.10    |0.03          |
+|Uniformity index |1.10^-6^       |6.10^-6^|3.10^-6^      |
+Table: Abstract of the heritability of the density, the ring width, parameters $b_0$, $b_1$, $b_2$ and the Variability index.
+
+Heritability is higher in each site separately than that of the combined sites for all traits.
+Wood density is the only character with strong heritability for each of the sites as well as for the combined sites.
+The annual ring width and the parameter $b_0$ have low heritability.
+When parameters $b_1$ and $b_2$, their heritability is very low.
+The uniformity index has almost zero heritability regardless of the study site.
+
+# 4 Discussion
+
+The data set showed a typical radial pattern of wood density already observed in spruce trees in other studies [@Alteyrac2006; @GARDINER2011; @Xiang2014].
+The trend is similar to that described as Type II density profile in @panshin1980.
+The high density during the first years of tree growth can be explained by the immature tracheids combined with the high proportion of compression wood.
+The density decreases to a certain level and begins to increase again.
+This increase is due to the fact that the width of the ring decreases from a certain age, so the proportion of initial decreases and that of the final wood [@Lachenbruch2011].
+The difference between the two sites at the parameter level can be explained by the fact that the climate between these sites is not the same which influences the variation of the density. 
+The modeling of individual profiles is suitable for the description of the wood density of our type II data.
+The choice of model allowed to take into account the variability of the density profiles of white spruce.
+@Xiang2014 applied a similar model in their study to describe the density profiles of black spruce with density profiles similar to those of white spruce.
+The low index of parameters $b_1$ and $b_2$ were most likely the large variability of the initial decline near the pith.
+The index of variability is different between the two sites, despite the fact that they are the same families of trees.
+This can be explained by the fact that both sites have different climatic conditions.
+It would be environmental.
+The fact that the index of variability is more under environmental control than genetic explains the lack of heritability of this index.
+It is not surprising that the variability index is under environmental control because it is calculated from model residuals.
+The model has made it possible to subtract the typical trend of the variation of the density which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control.
+Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.
+As for Spearman's rank correlation coefficient is not significant (p-value = 0.43; rho = 0.08) for the index of variability.
+This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the variability index is under environmental control.
+The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites taking into account the extreme variability of the different density profiles.
+Heritability of density is high at both sites.
+This strong density heritability is similar to previous studies in white spruce [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
+All of these studies reported values between 0.36 and 0.52.
+In contrast to density, annual ring width has low to moderate heritability in each site and also for combined sites.
+This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by the fact that growth is more influenced by environmental conditions than by genetics.
+The difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986].
+The parameters of the model, as for them, they have a low heritability either for each site separately or for the combined sites.
+The parameter b0, has a higher heritability than the other parameters but remains low.
+This can be explained by the fact that this parameter is very little influenced by cambial age.
+The index of variability has almost no heritability.
+It is not surprising that this trait has very low heritability because it is estimated from model residuals after subtracting the typical density profile.
+Our study includes limitations for identifying families that may have low variability in wood density as part in tree improvement programs.
+Future studies should focus on the resilience and variability associated with one-time events such as short-term water stress.
+In addition, the trees used for this study are relatively young, we can be interested in older trees.
 
 # Conclusion
+
 This study assessed the variability of wood density in white spruce.
@@
 
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
-
 \clearpage

### 2019-05-23 Article_AS.Rmd
Intro complétée

@@
 Therefore, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@Bonfils2004].
+It allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for the human use.
 \
@@
 
-During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+Conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
+Wood density is among the key traits as it is related mechanical strength and dimensional stability.
+Variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
 Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011, @Beaulieu2009].
 Wood quality has mainly two components: biologic and economic.
 The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
-Several traits were considered for wood quality assessment and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006] because this trait is important for both components of quality.
+Several traits were considered for wood quality assessment and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006] because this trait is important for both components of quality and strongly correlated with different mechanical and anatomical wood traits [@zobel1989].
 \
@@
 Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b, @Beaulieu2006] and this makes this trait attractive for genetic improvement.
+Moreover, Climate is a determining factor for the growth and physiological processes of trees.
+As a result, climatic conditions strongly affect the properties of wood.
+In fact, certain climatic factors such as drought can cause stress in trees and cause a variation in wood properties, including density, ring width and conductivity.
+In addition, long-term growth analyzes have shown that it may increase due to temperature, atmospheric CO2 content and nitrogen deposition [@becker1995; @huang2007; @zaehle2014].
+Knowing that density can be negatively correlated with growth in spruce [@zhang1995], it is essential to take this factor into account.
 
@@
 With the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005].
+It is necessary to anticipate the possible environmental variations because more important frequencies water stress or important precipitations are feared with the climatic changes [@IPCC2007; @pachauri2014].
+These environmental variations may lead to an increase or decrease in radial growth and thus cause greater variations in the density of the wood.
 In addition to selecting trees that are prone to produce wood of uniform density we must also ensure that these properties will persist in time with climatic change.
 Thus the selection must also considered the stability of the properties in time or, more accuralty, with climate change.
-\
+The heritability of density has been proven in several studies, but there is no evidence to suggest that density variability is hereditary.
+The heritability of this trait could qualify it for the selection criteria for future breeding programs for producing better quality wood.\
 
@@
 We ultimalty aim to select and reproduce trees families of interest that are both low-sensitivity to climate change and genetically prone to make homogenous wood.
+Thus, the most homogeneous families will be identified for reforestation and identify variables to consider in breeding of growing trees with less variable wood.
 
@@
 
-![Study sites](graph_table/Study.sites.png){width=320px}
+![Study sites](graph_table/Study.sites.png){width=500px; height=500px}
 

### 2019-05-27 Article_AS.Rmd
Modif intro

@@
   chunk_output_type: inline
-output: pdf_document
+output:
+  pdf_document: default
+  word_document: default
 bibliography: biblio.bibtex
@@
 \
-
 Improvement programs by genetic selection are in place for the most important commercial species [@DRF2000; @Mullin2011] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015].
@@
 In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@Nienstaedt1990].
-All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly accross almost all forested regions which represents a wide range of soils types and a climatic conditions (Canadian Council of Forest Ministers 2009).
+All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly accross almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009].
 \
@@
 During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
-Conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
-Wood density is among the key traits as it is related mechanical strength and dimensional stability.
-Variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
-Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
-For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011, @Beaulieu2009].
+Therefore, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
 Wood quality has mainly two components: biologic and economic.
 The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
-Several traits were considered for wood quality assessment and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006] because this trait is important for both components of quality and strongly correlated with different mechanical and anatomical wood traits [@zobel1989].
-\
-
-With regard to the biological component, in has been demonstrated [@Hietz2013, @osazuwa2014] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
+Several traits were considered for wood quality economic and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006] because this trait is important for both components of quality and strongly correlated with different mechanical (mechanical resistance and the dimensional stability) and anatomical wood traits [@zobel1989].
+Thus, variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
+Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011; @Beaulieu2009].
+With regard to the biological component, in has been demonstrated [@Hietz2013; @osazuwa2014] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
 For example a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
@@
 
-From these studies we understand that there might be a strong gain at selecting trees that are prone to produce wood of uniform density to get a resistant and higly valuable forest in the next decades.
-To reach this goal is is important to understant where do the variations of density come from.
+In view of the studies cited aboves, we understand that there might be a strong gain at selecting trees that are prone to produce wood of uniform density to get a resistant and higly valuable forest in the next decades.
+To reach this goal is important to understant where do the variations of density come from.
 In white spruce, there is a typical low frequency radial pattern of variation from pith to bark [@Moore2011].
 First a decrease of density during the first years of growth and then an increase from the beginning of the formation of mature wood [@panshin1980].
-In addition this radial variations there are hight frequency interannual variation [@Nicault2010].
+In addition, this radial variations there are hight frequency interannual variation [@Nicault2010].
 Contrary to the radial variation that is under internal control, the interannual variation of the density is influenced mainly by events events to the tree such as the edaphic and climatic conditions [@Nicault2010] that happen at high frequency.
@@
 
-Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b, @Beaulieu2006] and this makes this trait attractive for genetic improvement.
+Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] and this makes this trait attractive for genetic improvement.
 Moreover, Climate is a determining factor for the growth and physiological processes of trees.
@@
 In addition, long-term growth analyzes have shown that it may increase due to temperature, atmospheric CO2 content and nitrogen deposition [@becker1995; @huang2007; @zaehle2014].
-Knowing that density can be negatively correlated with growth in spruce [@zhang1995], it is essential to take this factor into account.
+Knowing that density can be negatively correlated with growth in spruce [@zhang1995], it is essential to take climat factor into account.
 
-However, the timeframe of such improved forest management being long [@Zobel1972] it is essential to take into account the current climate change context [@Gray2016].
+The timeframe of improved forest management being long [@Zobel1972] it is essential to take into account the current climate change context [@Gray2016].
 With the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005].
@@
 Thus the selection must also considered the stability of the properties in time or, more accuralty, with climate change.
-The heritability of density has been proven in several studies, but there is no evidence to suggest that density variability is hereditary.
+furthermore, heritability of density has been proven in several studies, but there is no evidence to suggest that density variability is hereditary.
 The heritability of this trait could qualify it for the selection criteria for future breeding programs for producing better quality wood.\

### 2019-05-29 Article_AS.Rmd
Fix wording

@@
 The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
-Several traits were considered for wood quality economic and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006] because this trait is important for both components of quality and strongly correlated with different mechanical (mechanical resistance and the dimensional stability) and anatomical wood traits [@zobel1989].
-Thus, variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
+Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
+Indeed this trait is important for both components of quality because it is strongly correlated with different mechanical and anatomical wood traits [@zobel1989].
+Variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
 Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].

### 2019-05-29 Article_AS.Rmd
MM

--- 2019-05-27 Article_AS.Rmd
@@
 This series of tests was established in 1999 by the Canadian Forest Service and is one of the three essential parts of the white spruce breeding program in Quebec.
-Two sites were selected for sampling: one near Saint-Casimir, Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather).
-Between nine and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees (fig 1).
-Trees were 15 years old at the time of sampling.
-Trees were assigned to row-plots of five trees/plot (2 m × 2 m spacing).
-Core samples were extracted at breast height (1.3 m above ground level) from the southern side of the trees to minimize the variation due to sample orientation.
+Two sites were selected for sampling: one near Saint-Casimir, Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather) (fig 1).
+Between 9 and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees.
 Tree height and diameter at breast height (DBH, 1.3 m above ground) were recorded in 2015 at the age of 15 years, and a wood increment core was extracted from the south facing side of each tree.
@@
 
-where RD is the mean ring density (kgm-3), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and e is the residual error term.
+where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and e is the residual error term.
 Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.
@@
 We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
-The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R2) [@Pape1999].
-The graphic description in figure 1 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).
+The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999].
+The graphic description in figure 2 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).
 
@@
 
-It can be observed in figure 1 that the parameters $b_0$ and $b_1$ influence the whole of the curve.
+It can be observed in figure 2 that the parameters $b_0$ and $b_1$ influence the whole of the curve.
 An increase of one of these parameters leads to that of the density on all rings.
@@
 
-## Determination of the wood density uniformity index
-The uniformity index is an index to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
-The residues of our model were used to obtain the wood density uniformity index, which was determined using an equation.
+## Determination of the wood density variability index
+The variability index is an index to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
+The residues of our model were used to obtain the wood density variability index, which was determined using an equation.
 We subtracted the density residues values of year $n-1$ from year $n$.
@@
 
-$$H = \sum | D_n - D_{n - 1} |$$
+$$V = \sum | D_n - D_{n - 1} |$$
 
-Where H is the uniformity index, $D_n$ is the wood density residues of year $N$ and  $D_n - 1$ is the tree density residues of year $N-1$.
+Where V is the variability index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the tree density residues of year $n-1$.
 
@@
 To describe the link between two variables by means of a monotonic function, we used Spearman’s rank correlation coefficient.
-We calculated this coefficient for the index of uniformity and the different parameters of the model.
-The elements for which we calculated Spearman’s rho are: the uniformity index for each of the tree families between the two study sites, the model residues for each of the families and the three parameters to estimate from the model.
+We calculated this coefficient for the index of variability and the different parameters of the model.
+The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites, the model residues for each of the families and the three parameters to estimate from the model.
 The cor.test function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho.

### 2019-05-30 Article_AS.Rmd
modif1 intro

@@
 
+### Context
 Canadian forest is an important part of the country's history, culture, natural environment and economy.
@@
 
+### Qualité du bois et ses composantes
 During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
@@
 The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
-Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
+With regard to the biological component, in has been demonstrated [@Hietz2013; @osazuwa2014] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
+For example a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
+With regard to the economical context, several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
 Indeed this trait is important for both components of quality because it is strongly correlated with different mechanical and anatomical wood traits [@zobel1989].
 Variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
-Consequently studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+In fact, wood density uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
+For exemple, a strong interannual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance.
+ It is therefore important from both a biological point of view and a end-use point of view to have a wood with a uniform density within the stem. \
+
+### Ce que l'on sait de l'uniformité du bois et son importance
+In view of the studies cited aboves, we understand that there might be a strong gain at selecting trees that are prone to produce wood of uniform density to get a resistant and higly valuable forest in the next decades.
+Consequently, studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
 For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011; @Beaulieu2009].
-With regard to the biological component, in has been demonstrated [@Hietz2013; @osazuwa2014] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
-For example a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
-With regard to the economical context, uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
-For exemple, a strong interannual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance.
-It is therefore important from both a biological point of view and a end-use point of view to have a wood with a uniform density within the stem.
-\
-
-In view of the studies cited aboves, we understand that there might be a strong gain at selecting trees that are prone to produce wood of uniform density to get a resistant and higly valuable forest in the next decades.
-To reach this goal is important to understant where do the variations of density come from.
+Hense, it is important to understant where do the variations of density come from in order to better integrated them into genetic selections.
 In white spruce, there is a typical low frequency radial pattern of variation from pith to bark [@Moore2011].
@@
 
+### Contrôle génétique et environnemental de la densité du bois
 Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] and this makes this trait attractive for genetic improvement.
-Moreover, Climate is a determining factor for the growth and physiological processes of trees.
-As a result, climatic conditions strongly affect the properties of wood.
-In fact, certain climatic factors such as drought can cause stress in trees and cause a variation in wood properties, including density, ring width and conductivity.
+Although variation in wood density is strongly controlled by genetics (30% - 60%) in white spruce, some of the variation is explained by other factors such as growth.
+Knowing that climate is a determining factor for the growth and physiological processes of trees.
+As a result, climatic conditions affect wood density.
+In fact, some climatic factors such as drought can cause stress in trees and cause a variation in wood properties, including density, ring width and conductivity.
 In addition, long-term growth analyzes have shown that it may increase due to temperature, atmospheric CO2 content and nitrogen deposition [@becker1995; @huang2007; @zaehle2014].
-Knowing that density can be negatively correlated with growth in spruce [@zhang1995], it is essential to take climat factor into account.
+Considering that density can be negatively correlated with growth in spruce [@zhang1995], it is essential to take climat factor into account.
 
-The timeframe of improved forest management being long [@Zobel1972] it is essential to take into account the current climate change context [@Gray2016].
-With the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005].
-It is necessary to anticipate the possible environmental variations because more important frequencies water stress or important precipitations are feared with the climatic changes [@IPCC2007; @pachauri2014].
-These environmental variations may lead to an increase or decrease in radial growth and thus cause greater variations in the density of the wood.
-In addition to selecting trees that are prone to produce wood of uniform density we must also ensure that these properties will persist in time with climatic change.
-Thus the selection must also considered the stability of the properties in time or, more accuralty, with climate change.
-furthermore, heritability of density has been proven in several studies, but there is no evidence to suggest that density variability is hereditary.
+### L'importance de la prise en compte du climat et de l'héritabilité des caractères dans les les programmes de selection génétique
+The timeframe of improved forest management being long [@Zobel1972] it is important to anticipate future changes in environmental conditions in these programs [@Gray2016].
+This will allow the establishment of plantations capable of adapting to the water stress and possible significant rainfall related to climate change because climate predictions predict more variations in the coming years [@IPCC2007; @pachauri2014].
+In fact, with the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005] hence, the importance of integrating these factors into genetic selection programs.
+However, to selecting trees that are prone to produce wood of uniform density we must also ensure that these properties will persist in time with climatic change.
+Thus the selection must also considered the stability of the properties in time or, more accurately, with climate change.
+Furthermore, heritability of density has been proven in several studies, but there is no evidence to suggest that density variability is hereditary.
 The heritability of this trait could qualify it for the selection criteria for future breeding programs for producing better quality wood.\
 
+### Le but de notre étude et ce que nous avons fait. 
 In this study we quantified and compared heteregoneity of wood density from trees of known parents in order to figure out if some individus produce more uniform wood than others.
@@
 
-![Study sites](graph_table/Study.sites.png){width=500px; height=500px}
+![Study sites](graph_table/Study.sites.png){width=400px; height=400px}
 
@@
 In table 1, we observe that the average value of the parameters differs between Asselin and Saint-Casimir.
-for the parameters b0 and b1, the average is higher in Saint-Casimir and for parameter b2 it is higher in Asselin.\
+for the parameters $b_0$ and $b_1$, the average is higher in Saint-Casimir and for parameter $b_2$ it is higher in Asselin.\
 
@@
 |                         |$b_2$     | -50.87|  14.87|  61.34|
-Table: Summary table of variability index
+Table: Summary table of parameters
 
@@
 
-![The ring density versus the number of annual rings of the samples for the Asselin site.The bold lines represent the model prediction](graph_table/Density_asselin.png){width=330px; height=330px}
+![The ring density versus the number of annual rings of the samples for the Asselin site.The bold lines represent the model prediction](graph_table/Density_asselin.png){width=330px; height=330px} \
 
-![The ring density versus the number of annual rings of the samples for the Saint-Casimir site.The bold lines represent the model prediction](graph_table/Density_casimir.png){width=330px; height=330px}
+![The ring density versus the number of annual rings of the samples for the Saint-Casimir site.The bold lines represent the model prediction](graph_table/Density_casimir.png){width=330px; height=330px} \
 
-Model predictions are similar to data trends across sites (fig3&4).
+Model predictions are similar to data trends across sites (fig 3&4).
 A decrease in density is observed in the first years of growth and then an increase in density from a certain cambial age.
@@
 The increase in density after the decrease of the first years of growth is greater in Saint-Casimir than in Asselin.
+However, the average density is very similar between the two sites.
 In spite of the interannual variations, the trees have the same general tendency and this one seems well described by the model which adapts to each of the sites.\
@@
 
+When comparing the index of variability between study sites, there is greater variability in Asselin than in Saint-Casimir.
+The maximum value of 1155.8 in Asselin, for example, is 746, which represents a huge difference.
+
 |Rank correlation  |  Rho| P-value|
@@
 |Variability index | 0.08|    0.43|
-|$b_0$             | 0.70|    0.00|
-|$b_1$             | 0.80|    0.00|
-|$b_2$             | 0.60|    0.00|
+|$b_0$             | 0.70|2.2^-16^|
+|$b_1$             | 0.80|2.2^-16^|
+|$b_2$             | 0.60|2.2^-16^|
 Table: Sperman’s Rang correlation
@@
 When parameters $b_1$ and $b_2$, their heritability is very low.
-The uniformity index has almost zero heritability regardless of the study site.
+The uniformity index has almost zero heritability regardless of the study site.\
 

### 2019-06-03 Article_AS.Rmd
Intro JR

@@
 
-### Context
 Canadian forest is an important part of the country's history, culture, natural environment and economy.
-But thousands of hectares of canadian forest are harvested each year and an additionnal loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
-To maintain ressource constant, it is therefore essential to compensate the sampling and the looses by reforesting.
+But thousands of hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
+To maintain resource constant, it is therefore essential to compensate the sampling and the looses by reforesting.
 The reforestation, when under human control, could ideally be made by using better trees than the former ones, i.e.
-trees that growth faster, that are more resistant to perturbation and that are of better stuctural quality.
+trees that growth faster, that are more resistant to perturbation and that are of better structural quality.
 To achieve this task, we must take into account the cambial activity that is governed by environmental conditions and genetics [@Dufour2010; @Rossi2012].
@@
 \
+
 Improvement programs by genetic selection are in place for the most important commercial species [@DRF2000; @Mullin2011] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015].
@@
 In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@Nienstaedt1990].
-All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly accross almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009].
+All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009].
 \
 
-### Qualité du bois et ses composantes
-During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of lossing ressource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
 Therefore, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
-Wood quality has mainly two components: biologic and economic.
-The biologic component is the resistance to mortality by vigor and the economic component is the value of the wood for end-use.
-With regard to the biological component, in has been demonstrated [@Hietz2013; @osazuwa2014] that trees with a more uniform wood densitiy within the stem are also the more resistant to mortality.
-For example a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
-With regard to the economical context, several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
-Indeed this trait is important for both components of quality because it is strongly correlated with different mechanical and anatomical wood traits [@zobel1989].
-Variation in density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
-In fact, wood density uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
-For exemple, a strong interannual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance.
- It is therefore important from both a biological point of view and a end-use point of view to have a wood with a uniform density within the stem. \
-
-### Ce que l'on sait de l'uniformité du bois et son importance
-In view of the studies cited aboves, we understand that there might be a strong gain at selecting trees that are prone to produce wood of uniform density to get a resistant and higly valuable forest in the next decades.
-Consequently, studies have been set up to understand the genetic sources of phenotypic variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
-For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011; @Beaulieu2009].
-Hense, it is important to understant where do the variations of density come from in order to better integrated them into genetic selections.
-In white spruce, there is a typical low frequency radial pattern of variation from pith to bark [@Moore2011].
-First a decrease of density during the first years of growth and then an increase from the beginning of the formation of mature wood [@panshin1980].
-In addition, this radial variations there are hight frequency interannual variation [@Nicault2010].
-Contrary to the radial variation that is under internal control, the interannual variation of the density is influenced mainly by events events to the tree such as the edaphic and climatic conditions [@Nicault2010] that happen at high frequency.
+Wood quality has a biologic and an economic component.
+Biologically, the resistance to mortality by vigour is important to ensure to harvest leaving trees.
+Economically the value of the wood for end-use is important for industries.
+Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
+More specifically the homogeneity of the density within the stem has an high importance.
+I has been demonstrated that trees with a more uniform wood density within the stem are also the more resistant to mortality [@Hietz2013; @osazuwa2014] because a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
+With regard to the economical context, variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
+In fact, wood density uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bio energy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
+For example, a strong inter-annual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance [@x,@y].
+It is therefore important from both a biological and an end-use points of view to have a wood with a uniform density within the stem.
 \
 
-### Contrôle génétique et environnemental de la densité du bois
-Unlike growth that is more controlled by environment than genetic and therefore has low to moderate heritability [@Ivkovich2002a], wood density in white spruce is under strong genetic control with moderate to high heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] and this makes this trait attractive for genetic improvement.
-Although variation in wood density is strongly controlled by genetics (30% - 60%) in white spruce, some of the variation is explained by other factors such as growth.
-Knowing that climate is a determining factor for the growth and physiological processes of trees.
-As a result, climatic conditions affect wood density.
-In fact, some climatic factors such as drought can cause stress in trees and cause a variation in wood properties, including density, ring width and conductivity.
-In addition, long-term growth analyzes have shown that it may increase due to temperature, atmospheric CO2 content and nitrogen deposition [@becker1995; @huang2007; @zaehle2014].
-Considering that density can be negatively correlated with growth in spruce [@zhang1995], it is essential to take climat factor into account.
+In view of the studies cited above, we understand that there might be a strong gain at selecting trees that are prone to produce wood with an homogeneous density to get a resistant and highly valuable forest in the next decades.
+Consequently, studies have been set up to understand the genetic sources of phenotype variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011; @Beaulieu2009].
+Hence, it is important to understand where do the variations of density come from in order to better integrated them into genetic selections.
+In white spruce, there is a typical low frequency radial pattern of variation from pith to bark [@Moore2011] with a density that decreases during the first years of growth and then an increasing from the beginning of the formation of mature wood [@panshin1980].
+These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
+This makes this trait attractive for genetic improvement.
+In addition to these low-frequency variations, there are high frequency inter-annual variations [@Nicault2010] that are influenced mainly by events to the tree such as the edaphic and climatic conditions [@Nicault2010] that happen at high frequency.
+These variations are not under genetic control but under climate control meaning that climatic conditions also affect wood density.
+Some climatic factors such as drought can cause stress in trees and cause a variation in wood properties, including density, ring width and conductivity.
+Long-term growth analyses have shown that density may even increases with the temperature, the atmospheric CO~2~ content and the nitrogen deposition [@becker1995; @huang2007; @zaehle2014].
+Wood density can also be negatively correlated with growth.
+[@zhang1995].
+Therefore, with the aim of producing wood of homogeneous density it is essential to take climate factors into account in addition to genetic.
 
 ### L'importance de la prise en compte du climat et de l'héritabilité des caractères dans les les programmes de selection génétique
-The timeframe of improved forest management being long [@Zobel1972] it is important to anticipate future changes in environmental conditions in these programs [@Gray2016].
+<!-- Ca part en couille il n'y a plus de logique. Quel message veux tu faire passer ? Tu te tires une balle dans le pied en disant l'inverse de que tu as dit avant ! --> 
+The time-frame of improved forest management being long [@Zobel1972] it is important to anticipate future changes in environmental conditions in these programs [@Gray2016].
 This will allow the establishment of plantations capable of adapting to the water stress and possible significant rainfall related to climate change because climate predictions predict more variations in the coming years [@IPCC2007; @pachauri2014].
 In fact, with the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005] hence, the importance of integrating these factors into genetic selection programs.
-However, to selecting trees that are prone to produce wood of uniform density we must also ensure that these properties will persist in time with climatic change.
+However, to select trees that are prone to produce wood of homogeneous density we must also ensure that these properties will persist in time with climatic change.
 Thus the selection must also considered the stability of the properties in time or, more accurately, with climate change.
@@
 ### Le but de notre étude et ce que nous avons fait. 
-In this study we quantified and compared heteregoneity of wood density from trees of known parents in order to figure out if some individus produce more uniform wood than others.
-Knowing the parents of each tree we tried to characterize the genetic contribution to the homogeneity.
-By comparing two growth sites we tried to characterize the environmental contribution to homogeity.
-We ultimalty aim to select and reproduce trees families of interest that are both low-sensitivity to climate change and genetically prone to make homogenous wood.
+In this study we quantified and compared heterogeneity of wood density from trees of known parents in order to figure out if some individuals produce more heterogeneous wood than others.
+Knowing the parents of each tree we tried to characterize the genetic contribution to the heterogeneity by comparing two growth sites we tried to characterize the environmental contribution to homogeneity.
+We ultimately aim to select and reproduce trees families of interest that are both low-sensitivity to climate change and genetically prone to make homogeneous wood.
 Thus, the most homogeneous families will be identified for reforestation and identify variables to consider in breeding of growing trees with less variable wood.
@@
 Increasing the value of the parameters $b_0$ and $b_1$ results in an increase in the density at the level of all the rings.
-The increase in the parameter $b_2$ causes a sharp rise in density after the decrease in the first rings.](graph_table/parameters.pdf)
+The increase in the parameter $b_2$ causes a sharp rise in density after the decrease in the first rings.](graph_table/parameters.png)
 

### 2019-06-03 Article_AS.Rmd
Discussion1

--- 2019-05-30 Article_AS.Rmd
@@
 This increase is due to the fact that the width of the ring decreases from a certain age, so the proportion of initial decreases and that of the final wood [@Lachenbruch2011].
-The difference between the two sites at the parameter level can be explained by the fact that the climate between these sites is not the same which influences the variation of the density. 
 The modeling of individual profiles is suitable for the description of the wood density of our type II data.
@@
 The low index of parameters $b_1$ and $b_2$ were most likely the large variability of the initial decline near the pith.
-The index of variability is different between the two sites, despite the fact that they are the same families of trees.
-This can be explained by the fact that both sites have different climatic conditions.
-It would be environmental.
+Although the variaton of density follows a Type II profile, it remains slightly different between the two sites.
+This difference can be explained by the difference in pedoclimatic cinditions between sites.
+Although, the soil type is different between sites and the climate is colder in Asselin compared to Saint-Casimir.
+At the level of the model parameters, the sites are also different and this difference can be explained by the profiles of densities that are not the same.\
+
+The index of variability is different between two sites, despite the fact that they are the same families of trees.
+Since the variability index is calculated from the model residuals, the difference can be explained by the fact that the variability that is related to the internal control of the trees has been subtracted from the model and that the remaining one is more controlled by external phenomena related to the study sites.
 The fact that the index of variability is more under environmental control than genetic explains the lack of heritability of this index.
-It is not surprising that the variability index is under environmental control because it is calculated from model residuals.
+It is not surprising that the variability index is under environmental control because density is mainly controlled by genetics but partly also by growth that is mainly controlled by environmental conditions.
 The model has made it possible to subtract the typical trend of the variation of the density which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control.
-Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.
-As for Spearman's rank correlation coefficient is not significant (p-value = 0.43; rho = 0.08) for the index of variability.
+Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.\
+
+The index of variability being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index.
 This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the variability index is under environmental control.
-The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites taking into account the extreme variability of the different density profiles.
+The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites taking into account the extreme variability of the different density profiles.\
+
 Heritability of density is high at both sites.
@@
 In contrast to density, annual ring width has low to moderate heritability in each site and also for combined sites.
-This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by the fact that growth is more influenced by environmental conditions than by genetics.
-The difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986].
+This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by The difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986], this happens when conditions from one environment to another do not have the same effect on the genotype.
+This result suggests that, if trees are to be selected for reforestation as a result of this study, the territory be divided into relatively homogeneous subunits, within which the recommended sources will be able to keep their superiority and can be used without risk of maladjustment [@beaulieu1989stabilite].
+
 The parameters of the model, as for them, they have a low heritability either for each site separately or for the combined sites.
-The parameter b0, has a higher heritability than the other parameters but remains low.
+The parameter $b_0$, has a higher heritability than the other parameters but remains low.
 This can be explained by the fact that this parameter is very little influenced by cambial age.
+In fact, $b_0$ is the ordinate at the origin and thus little influenced by the sites (environmental conditions).
 The index of variability has almost no heritability.

### 2019-06-05 Article_AS.Rmd
Intro version1

@@
 author: "André Dogninema Soro, Jean-Romain Roussel, Alexis Achim"
-editor_options:
-  chunk_output_type: inline
 output: 
@@
 bibliography: biblio.bibtex
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 But thousands of hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
-To maintain resource constant, it is therefore essential to compensate the sampling and the looses by reforesting.
+To maintain resource constant, it is therefore essential to compensate the harvest and the looses by reforesting.
 The reforestation, when under human control, could ideally be made by using better trees than the former ones, i.e.
@@
 More specifically the homogeneity of the density within the stem has an high importance.
-I has been demonstrated that trees with a more uniform wood density within the stem are also the more resistant to mortality [@Hietz2013; @osazuwa2014] because a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
+Its has been demonstrated that trees with a more uniform wood density within the stem are also the more resistant to mortality [@Hietz2013; @osazuwa2014] because a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
 With regard to the economical context, variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
 In fact, wood density uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bio energy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
-For example, a strong inter-annual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance [@x,@y].
+For example, a wide inter-annual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance [@x,@y].
 It is therefore important from both a biological and an end-use points of view to have a wood with a uniform density within the stem.
@@
 
-### L'importance de la prise en compte du climat et de l'héritabilité des caractères dans les les programmes de selection génétique
-<!-- Ca part en couille il n'y a plus de logique. Quel message veux tu faire passer ? Tu te tires une balle dans le pied en disant l'inverse de que tu as dit avant ! --> 
-The time-frame of improved forest management being long [@Zobel1972] it is important to anticipate future changes in environmental conditions in these programs [@Gray2016].
-This will allow the establishment of plantations capable of adapting to the water stress and possible significant rainfall related to climate change because climate predictions predict more variations in the coming years [@IPCC2007; @pachauri2014].
-In fact, with the increasing temperature, the trees are expected to growth faster and thus being of lower quality because of lower density [@corriveau1987; @DeBell1994; @Lindstrom1996; @Dutilleul1998; @Bouriaud2005] hence, the importance of integrating these factors into genetic selection programs.
-However, to select trees that are prone to produce wood of homogeneous density we must also ensure that these properties will persist in time with climatic change.
-Thus the selection must also considered the stability of the properties in time or, more accurately, with climate change.
-Furthermore, heritability of density has been proven in several studies, but there is no evidence to suggest that density variability is hereditary.
-The heritability of this trait could qualify it for the selection criteria for future breeding programs for producing better quality wood.\
+The impact of climate is more perceptible on the properties of wood when there are wide variations of it such as drought, high temperatures or significant rainfall.
+The ability of trees to tolerate these changes in a tree species or within a species is often related to the interaction between genotype and environment.
+For example, studies have shown a low sensibility to genotype-related drought [@ferrell1966; @teskey1987].
+Consequently, the sensitivity of trees to climate such as inter-annual variability in density can be controlled by genetics.
+It is therefore important to use trees that are less sensitive to climate variations because of their genome for breeding programs.
 
-### Le but de notre étude et ce que nous avons fait. 
 In this study we quantified and compared heterogeneity of wood density from trees of known parents in order to figure out if some individuals produce more heterogeneous wood than others.
 Knowing the parents of each tree we tried to characterize the genetic contribution to the heterogeneity by comparing two growth sites we tried to characterize the environmental contribution to homogeneity.
-We ultimately aim to select and reproduce trees families of interest that are both low-sensitivity to climate change and genetically prone to make homogeneous wood.
+We ultimately aim to select and reproduce white spruce trees families of interest that are both low-sensitivity to climate change and genetically prone to make homogeneous wood.
 Thus, the most homogeneous families will be identified for reforestation and identify variables to consider in breeding of growing trees with less variable wood.
@@
 Tree height and diameter at breast height (DBH, 1.3 m above ground) were recorded in 2015 at the age of 15 years, and a wood increment core was extracted from the south facing side of each tree.
-Cores were stored in a freezer, conditioned to 7% moisture and cut to 1.68 mm thickness prior to X-ray densitometry analyses (Quintek Measurement Systems, TN).
+Cores were stored in a freezer, conditioned to 7% moisture and cut to 1.68 mm thickness prior to X-ray densitometry analyses (Quintek Measurement Systems, TN) at a resolution of 0.0254 mm, and the data were reported as relative wood density on a dry weight basis.
 Wood density was calculated as a ring area weighted mean from recorded pith to bark wood density profiles.
 
-![Study sites](graph_table/Study.sites.png){width=400px; height=400px}
+![Map of Quebec showing the location of the two study sites where the data were collected](graph_table/Study.sites.png){width=300px; height=400px}
 
-Growth ring density and ring width measurements were taken from pith-to-bark, using an X-ray densitometer (Quintek Measurement Systems, TN) at a resolution of 0.0254 mm, and the data were reported as relative wood density on a dry weight basis.
 To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983].
+thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be modified.
 
@@
 
-where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and e is the residual error term.
+where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term.
 Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.
@@
 It can be observed in figure 2 that the parameters $b_0$ and $b_1$ influence the whole of the curve.
-An increase of one of these parameters leads to that of the density on all rings.
+An increase of one of these parameters leads to that of the density on all the ring.
 Unlike parameter $b_0$ and $b_1$, parameter $b_2$ seems to be influenced by some of the density.
-A low value of this parameter causes a sharp increase of the density starting from the third ring and when this parameter increases a decrease of the density is observed at this level until the curve is similar to the cases of the parameters $b_0$ and $b_1$.
-Beyond the value 1, the parameter $b_2$ does not influence the variation of the density.\
+A low value of this parameter causes a sharp increase of the density starting from the third growth year and when this parameter increases a decrease of the density is observed at this level until the curve is similar to the cases of the parameters $b_0$ and $b_1$.
+\
 
 ## Determination of the wood density variability index
-The variability index is an index to determine the variation in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
+The variability index is an index to determine the uniformity in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
 The residues of our model were used to obtain the wood density variability index, which was determined using an equation.
@@
 
-Where V is the variability index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the tree density residues of year $n-1$.
+Where V is the variability index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
 
@@
 In table 1, we observe that the average value of the parameters differs between Asselin and Saint-Casimir.
-for the parameters $b_0$ and $b_1$, the average is higher in Saint-Casimir and for parameter $b_2$ it is higher in Asselin.\
+For the parameters $b_0$ and $b_1$, the average is higher in Saint-Casimir and for parameter $b_2$ it is higher in Asselin.\
 

### 2019-06-12 Article_AS.Rmd
Intro apres commentaires generaux de AA

@@
 title: "On the eletrodynamic of moving bodies"
-author: "André Dogninema Soro, Jean-Romain Roussel, Alexis Achim"
+author: "André D. Soro, Patrick Lenz, Jean Bousquet, Mariana Hassegawa, Jean-Romain Roussel, Alexis Achim"
 output: 
@@
 
-Canadian forest is an important part of the country's history, culture, natural environment and economy.
-But thousands of hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
-To maintain resource constant, it is therefore essential to compensate the harvest and the looses by reforesting.
+Thousands hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
+It is important to increase tree growth to meet the growing demand of forest industry.
 The reforestation, when under human control, could ideally be made by using better trees than the former ones, i.e.
@@
 Therefore, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@Bonfils2004].
-It allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for the human use.
+It allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for the human use [@zobel1984; @DRF2000; @nanson2004].
 \
@@
 
-During the first half of the 20th century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+During the first half of the 20^th^ century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
 Therefore, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
-Wood quality has a biologic and an economic component.
+Stem vigor (biologic component) and wood quality (economic component) are two desirable traits that are taken into account in genetic selection.
 Biologically, the resistance to mortality by vigour is important to ensure to harvest leaving trees.
@@
 Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
-More specifically the homogeneity of the density within the stem has an high importance.
-Its has been demonstrated that trees with a more uniform wood density within the stem are also the more resistant to mortality [@Hietz2013; @osazuwa2014] because a uniform density gives a greater mechanical resistance to the wind and therefore a lower probability of mortality during a climatic event such as a storm.
-With regard to the economical context, variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
-In fact, wood density uniformity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bio energy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
-For example, a wide inter-annual variation of the density of the wood leads to a decrease of the mechanical resistance while a relatively uniform density produce wood with better mechanical performance [@x,@y].
-It is therefore important from both a biological and an end-use points of view to have a wood with a uniform density within the stem.
-\
+More specifically the homogeneity of density and rings width within the stem has an high importance for stem vigor and wood quality because these two traits are related to it.
+Homogeneity refers here to a small interannual variation in wood density and rings width.\
 
@@
 Consequently, studies have been set up to understand the genetic sources of phenotype variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
-For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Mullin2011; @Beaulieu2009].
+For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Beaulieu2009; @Mullin2011].
 Hence, it is important to understand where do the variations of density come from in order to better integrated them into genetic selections.
@@
 This makes this trait attractive for genetic improvement.
-In addition to these low-frequency variations, there are high frequency inter-annual variations [@Nicault2010] that are influenced mainly by events to the tree such as the edaphic and climatic conditions [@Nicault2010] that happen at high frequency.
+In addition to these low-frequency variations, there are high frequency inter-annual variations [@Nicault2010] that are influenced mainly by events to the tree such as the soil water availability and climatic conditions [@Nicault2010] that happen at high frequency.
 These variations are not under genetic control but under climate control meaning that climatic conditions also affect wood density.
@@
 [@zhang1995].
-Therefore, with the aim of producing wood of homogeneous density it is essential to take climate factors into account in addition to genetic.
+Therefore, with the aim of producing wood of homogeneous density it is essential to take climate factors into account in addition to genetic.\
 
-The impact of climate is more perceptible on the properties of wood when there are wide variations of it such as drought, high temperatures or significant rainfall.
+ In fact, the density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a small variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008].
+For example, @suarez2004 found that at the tree scale, individuals with variable growth were more prone to die of drought than more regularly growing trees.
+These studies have shown that the steady growth and therefore the small variation in the density of the wood allows the tree to be more vigorous.
+With regard to the economical context, variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
+Indeed, wood density homogeneity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bio energy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
+For example, the low variation in wood density is appreciated for sawnwood because sawyers place a high value on the homogeneity of wood density [Hassegawa et al (pending approval); @X].
+It is therefore important from both a biological and an end-use points of view to have a wood with a uniform density within the stem.
+\
+
+A significant variation of the climate can cause a variation of the properties of the wood over several years.
+For example, several studies have shown that after the great drought of 2003 in Europe, the growth of forests has been affected over several years [@ciais2005; @granier2007; @reichstein2007].
 The ability of trees to tolerate these changes in a tree species or within a species is often related to the interaction between genotype and environment.
-For example, studies have shown a low sensibility to genotype-related drought [@ferrell1966; @teskey1987].
-Consequently, the sensitivity of trees to climate such as inter-annual variability in density can be controlled by genetics.
-It is therefore important to use trees that are less sensitive to climate variations because of their genome for breeding programs.
+For example, studies have suggested a possible link between low sensitivity to drought and genotype [@ferrell1966; @teskey1987].
+This link is poorly controlled and therefore poorly known.
+It is in context that this study fits to try to advance the knowledge in this direction.\
 
@@
 This series of tests was established in 1999 by the Canadian Forest Service and is one of the three essential parts of the white spruce breeding program in Quebec.
-Two sites were selected for sampling: one near Saint-Casimir, Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather) (fig 1).
+Two sites were selected for sampling: one near Porneuf (Saint-Casimir), Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather) (fig 1).
 Between 9 and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees.
@@
 To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983].
-thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be modified.
+Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be modified.
+Any incomplete or false rings were excluded from the dataset.
 
@@
 A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce.
-Considering the hierarchical structure of the data, with annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
+Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
 This type of model also considers the autocorrelation between the observations [@Lindstrom1990].
@@
 To describe the link between two variables by means of a monotonic function, we used Spearman’s rank correlation coefficient.
-We calculated this coefficient for the index of variability and the different parameters of the model.
-The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites, the model residues for each of the families and the three parameters to estimate from the model.
-The cor.test function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho.
+We calculated this coefficient for the variability index and the different parameters of the model.
+The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites and the three parameters to estimate from the model.
+The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho.
 We used a 95 percent confidence interval.
@@
 ## Quantitative genetic analyses
-Variance and covariance for genetics were estimated using mixed models [@Lenz2013] of the R software. The model used is as follows:
+Variance and covariance for genetics were estimated using mixed models [@Lenz2013] of the R software. The equation used is as follows:
 
@@
 
-where $yijkl$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
+where $y_{ijkl}$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
 In a full-sib design with random mating, members of the same family are presumed to share half of their genes [@white2007]. Therefore, individual heritability was obtained as follows:
 
- $$h_2= 2 \frac{V_f}{V_T}$$
+ $$h_2= 2 \frac{\sigma^2_f}{\sigma^2_T}$$
 
-Where $h_2$ is individual heritability, $V_f$ is family variance, $V_T$ is total variance.
+Where $h_2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_T$ is total variance.
 ASReml3 package [@Butler2018] in the R statistical programming environment [@R2018] was used to estimate trait genetic parameters based on the residual maximum likelihood.
@@
 
-The model we have developed has made it possible to optimize the density of the wood for each study site.
+The model we have developed has made it possible to optimize the density of the wood for each core.
 In table 1, we observe that the average value of the parameters differs between Asselin and Saint-Casimir.
@@
 
-When considering all the cores of trees from the pith-to-bark, there is initially a rapid decrease in density in the first five years.
+When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings.
 It goes from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to kg/m^3^ for the Asselin site.
@@
 |$b_2$            |0.04           |0.10    |0.03          |
-|Uniformity index |1.10^-6^       |6.10^-6^|3.10^-6^      |
+|Variability index |1.10^-6^       |6.10^-6^|3.10^-6^      |
 Table: Abstract of the heritability of the density, the ring width, parameters $b_0$, $b_1$, $b_2$ and the Variability index.
@@
 The density decreases to a certain level and begins to increase again.
-This increase is due to the fact that the width of the ring decreases from a certain age, so the proportion of initial decreases and that of the final wood [@Lachenbruch2011].
+This rapid decline in density is due to the significant growth of annual rings and the following increase is explained by the fact that the ring width  decreases from a certain age, because of the proportion of earlywood and latewood by ring.[@Lachenbruch2011].\
+
 The modeling of individual profiles is suitable for the description of the wood density of our type II data.
@@
 @Xiang2014 applied a similar model in their study to describe the density profiles of black spruce with density profiles similar to those of white spruce.
-The low index of parameters $b_1$ and $b_2$ were most likely the large variability of the initial decline near the pith.
+\
+
 Although the variaton of density follows a Type II profile, it remains slightly different between the two sites.
-This difference can be explained by the difference in pedoclimatic cinditions between sites.
-Although, the soil type is different between sites and the climate is colder in Asselin compared to Saint-Casimir.
+This difference can be explained by the difference in pedoclimatic conditions between sites.
+In fact, the soil type is different between sites and the climate is colder in Asselin which is further south compared to Saint-Casimir.
 At the level of the model parameters, the sites are also different and this difference can be explained by the profiles of densities that are not the same.\

### 2019-06-21 Article_AS.Rmd
M&M

@@
 Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008].
-In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@Nienstaedt1990].
+In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@nienstaedt1990].
 All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009].
@@
 ## Sampling
-The dataset used in this study came from genetic selection for a series of tests according to provenances [@Beaulieu1996].
-The improvement program for this region was led by Quebec’s Ministry of Forests, Wildlife and Parcs, Canada.
-Experimental plantings were initiated from white spruce seedlings grown in nurseries and planted at two years of age at a regular spacing between trees of 2 m in all directions.
-This series of tests was established in 1999 by the Canadian Forest Service and is one of the three essential parts of the white spruce breeding program in Quebec.
-Two sites were selected for sampling: one near Porneuf (Saint-Casimir), Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather) (fig 1).
-Between 9 and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees.
-Tree height and diameter at breast height (DBH, 1.3 m above ground) were recorded in 2015 at the age of 15 years, and a wood increment core was extracted from the south facing side of each tree.
-Cores were stored in a freezer, conditioned to 7% moisture and cut to 1.68 mm thickness prior to X-ray densitometry analyses (Quintek Measurement Systems, TN) at a resolution of 0.0254 mm, and the data were reported as relative wood density on a dry weight basis.
-Wood density was calculated as a ring area weighted mean from recorded pith to bark wood density profiles.
+
+The material used was sampled from an experimental provenance-progeny test plantation designed to assess the genetic merit of first generation selections, which had been subdivided into breeding groups and delineated by provenance or geographic region.
+The details of the setting up of the experiment are described in @beaulieu1996.
+Crosses were made according to a partial diallel mating design within each of the six different breeding groups or sublines [@Burdon1983], which were designed to limit future inbreeding to within group.
+Each parent tree has been used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups.
+Genetic testing was conducted in 1999 on 2-year-old nursery seedlings near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1).
+Trees were assigned to plots in rows of five trees / plot (spacing of 2 m × 2 m) [@beaulieu2014].
+In this study, 93 full-brother families for a total of 2365 trees.
+The sampled trees were drawn from 18 blocks over the two sites and 9–14 individuals were sampled per full-sib cross.
+Trees used in this study are exclusively full-brother families.\
+
+|Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |
+|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|
+|Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |
+|Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |
+Table: Location and climatic conditions* (over 1981–2010) of the two plantation sites.
+
+*LAT, latitude; LON, longitude; ELV, elevation; MAT, mean annual temperature; TAP, total annual precipitation.
+
+Wood cores were extracted à une épaisseur de 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground).
+Tree cores are extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself.
+These cores were used to determine ring width and wood density.
+Density was determined by scanning incremental cores from bark to bark along the radial face at a resolution of 25 $\mu$m in the QTRS-01X ring analyzer (Quintek Measurement Systems Inc., Knoxville, TN, USA).
+A weighted mean density line was obtained by summing the product between the annual density measurements and the ratio of the respective radial areas to the total radial area.
+Wood cores were conditioned to about 7% moisture before scanning.
+Thus, the determination of the density of the wood makes it possible to calculate a quality index of it.
+From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines.
+Ring width determined through information of densitometry profiles.\
 
@@
 
+As part of this study, it is essential to accurately identify the year of formation of the annual rings measured.
+To this end, we used the technique of Cross-dating which consists in obtaining the concordance of the number of rings and of their width pattern to avoid counting errors or those caused by missing rings or by false annual rings.
+Crossdating is the most basic principle of dendrochronology.
+Cross-dating is a technique that ensures each individual tree ring is assigned its exact year of formation.
+This is accomplished by matching patterns of wide and narrow rings between cores from the same tree, and between trees from different locations.
 To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983].
-Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be modified.
+Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be corrected.
 Any incomplete or false rings were excluded from the dataset.
@@
 ## Model Development
-A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce.
-Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
-This type of model also considers the autocorrelation between the observations [@Lindstrom1990].
-Thus, the estimates of the parameters can vary randomly around the average values of each level of the data structure (site, block, tree, core) [@Auty2013].
+A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations.
+Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018] (Eq.3).
+Previous studies have used this method to describe the density profile of different spruce species that have a similar profile[@franceschini2010; @gardiner2011].
+However, when modeling density profiles we found that the variation between individual discs follows different patterns from the marrow to the bark between tree.
+We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
+To work around this problem, we chose to model each of the profiles separately.
+We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
+The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999].
 As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores.
@@
 Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.
-We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
-We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
-The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999].
 The graphic description in figure 2 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).
@@
 
+This modeling of the density profiles then made it possible to subtract the estimated profile which represents the low frequency signal share described previously in the introduction.
+Thus, only the part of the high-frequency signal that interests us remains to evaluate the variability of the density related to factors external to the tree.\
+
 ## Determination of the wood density variability index
+After subtraction of the typical density, it is observed that the residues remain variable.
+Knowing that the low frequency signal is subtracted, this variation can only be due to the high frequency signal which is the interannual variation.
+As explained earlier in the introduction, low interannual variability is an important criterion for wood quality.
+In order to identify tree families with less interannual variability in density, we have created a viability index.
 The variability index is an index to determine the uniformity in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
@@
 ## Spearman’s rank correlation coefficient (Spearman’s rho) 
-To describe the link between two variables by means of a monotonic function, we used Spearman’s rank correlation coefficient.
+Spearman’s rank correlation coefficient is a correlation coefficient to describe the link between two variables by means of a monotonic function.
+We therefore used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites.
 We calculated this coefficient for the variability index and the different parameters of the model.
@@
 We used a 95 percent confidence interval.
+
+$$rho = \frac{cov(rg_X,rg_Y)}{\sigma_{rg_X} \times\sigma_{rg_Y}}$$
+Where $rho$ is a nonparametric measure of rank correlation, $cov(rg_X,rg_Y)$ is the covariance of the rank variables.
+$\sigma_{rg_X}$ and $\sigma_{rg_Y}$ are the standard deviations of the rank variables.
 
@@
 
+![Pith-to-bark radial pattern of breast height density by ring.
+Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px}.\
+
+![Pith-to-bark radial pattern of breast height density by ring.
+Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px}.
+\
+
 When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings.
 It goes from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to kg/m^3^ for the Asselin site.
-Then, this decrease was followed by a slight increase until the twelfth year.\
-
-![The ring density versus the number of annual rings of the samples for the Asselin site.The bold lines represent the model prediction](graph_table/Density_asselin.png){width=330px; height=330px} \
-
-![The ring density versus the number of annual rings of the samples for the Saint-Casimir site.The bold lines represent the model prediction](graph_table/Density_casimir.png){width=330px; height=330px} \
+Then, this decrease was followed by a slight increase until the twelfth year.
+\
 
@@
 |:-----------------|----:|-------:|
+|Density residus   | 0.57|8.7^-09^|
 |Variability index | 0.08|    0.43|
@@
 
-The data set showed a typical radial pattern of wood density already observed in spruce trees in other studies [@Alteyrac2006; @GARDINER2011; @Xiang2014].
-The trend is similar to that described as Type II density profile in @panshin1980.
-The high density during the first years of tree growth can be explained by the immature tracheids combined with the high proportion of compression wood.
-The density decreases to a certain level and begins to increase again.
-This rapid decline in density is due to the significant growth of annual rings and the following increase is explained by the fact that the ring width  decreases from a certain age, because of the proportion of earlywood and latewood by ring.[@Lachenbruch2011].\
+The dataset showed a typical radial pattern of wood density already observed in spruce trees in other researchers [@Alteyrac2006; @GARDINER2011; @Xiang2014] with the density initially decreasing outwards from the pith before slowly increasing towards an asymptotic value.
+This pattern corresponds to the type II pith-to-bark profile described by @panshin1980.
+The high density of wood adjacent to the pith has been attributed to high earlywood density and low latewood proportion [@alteyrac2007; @koubaa2007].
+In addition, in the first years of a tree's life, it produces "flexure wood" that is high density [@telewski1989].
+This type of wood is formed because the low rigidity of the young trees that allows them to withstand environmental constraints without breaking.
+The proportion of this wood decreases rapidly in the first few rings.
+The increase in density that follows the decline of the latter is related to the proportion of latewood that increase in mature wood [@clark2007].
+Similar hypotheses have been proposed to explain radial variation in wood properties by @Lachenbruch2011.
+However, this radial pattern does not take into account the interannual variability of density.
+\
 
 The modeling of individual profiles is suitable for the description of the wood density of our type II data.
-The choice of model allowed to take into account the variability of the density profiles of white spruce.
-@Xiang2014 applied a similar model in their study to describe the density profiles of black spruce with density profiles similar to those of white spruce.
-\
+The prediction of the model on all of the data that can be seen in Figures 3 and 4 is an example of the modeling that is done for individual profiles.
+The choice of modeling the variation of the density of each of the trees made it possible to adapt the model to the variations between the trees.
+The fact that the model equation is composed of two components has given sufficient flexibility to properly describe type II density.
+This type of equation has been used previously in the study of @Xiang2014, where the authors used a three-component equation.
+In their case, the third component was used to describe a decline near the bark, which is not the case with what is observed in our study.\
 
-Although the variaton of density follows a Type II profile, it remains slightly different between the two sites.
+Although the density variation to a similar profile, it remains slightly different between the two sites.
 This difference can be explained by the difference in pedoclimatic conditions between sites.
 In fact, the soil type is different between sites and the climate is colder in Asselin which is further south compared to Saint-Casimir.
-At the level of the model parameters, the sites are also different and this difference can be explained by the profiles of densities that are not the same.\
+In addition, the young age (15 years) of our trees could have an impact on the difference in density profiles between study sites, because it was observed that in white spruce, the environmental control being higher than the genetic control the first years of life of the tree [@Corriveau1991], the conditions of the site thus played an important role.
+Indeed, the observation of the parameter $b_2$ of model equation allows to see that the rise of the density profile is related to the second term of the equation, and a great value of age cambial causes a greater rise of the profile of density.
+It can be hypothesised that the effect of the site on density profiles would have been smaller in an older stand.
+This result is in accordance with observations in white spruce  [@Corriveau1991] and black spruce [@Xiang2014].
+At the level of the model parameters, the sites are also different and this difference explained by the profiles of density that are not the same.\
 
-The index of variability is different between two sites, despite the fact that they are the same families of trees.
+Like the density profiles, variability index is different between the two sites, despite the fact that they are the same families of trees.
 Since the variability index is calculated from the model residuals, the difference can be explained by the fact that the variability that is related to the internal control of the trees has been subtracted from the model and that the remaining one is more controlled by external phenomena related to the study sites.
-The fact that the index of variability is more under environmental control than genetic explains the lack of heritability of this index.
+Variability index being more under environmental control than genetic, this explain the lack of heritability of this index.
 It is not surprising that the variability index is under environmental control because density is mainly controlled by genetics but partly also by growth that is mainly controlled by environmental conditions.
-The model has made it possible to subtract the typical trend of the variation of the density which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control.
+The model has made it possible to subtract the typical pattern of the density variation which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control.
 Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.\
 
-The index of variability being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index.
+Variability index being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index.
 This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the variability index is under environmental control.
@@
 
-Heritability of density is high at both sites.
-This strong density heritability is similar to previous studies in white spruce [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
-All of these studies reported values between 0.36 and 0.52.
+With regard to heritability of density, it is high at both sites.
+This strong density heritability is similar to previous studies in white spruce [Nelson 1963 in @mothe1983; @Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
+All of these studies reported values between 0.36 and 0.80.
 In contrast to density, annual ring width has low to moderate heritability in each site and also for combined sites.
-This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by The difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986], this happens when conditions from one environment to another do not have the same effect on the genotype.
+This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by the difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986], this happens when conditions from one environment to another do not have the same effect on the genotype.
 This result suggests that, if trees are to be selected for reforestation as a result of this study, the territory be divided into relatively homogeneous subunits, within which the recommended sources will be able to keep their superiority and can be used without risk of maladjustment [@beaulieu1989stabilite].

### 2019-07-03 Article_AS.Rmd
intro version 2

@@
 Therefore, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@Bonfils2004].
-It allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for the human use [@zobel1984; @DRF2000; @nanson2004].
+In forestry, genetic selection is a process allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for human use [@zobel1984; @DRF2000; @nanson2004].
 \
 
-Improvement programs by genetic selection are in place for the most important commercial species [@DRF2000; @Mullin2011] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015].
+During the first half of the 20^th^ century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
+Hence, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
+For example, improvement programs by genetic selection are in place for white spruce *(Picea glauca [Moench] Voss)* [@DRF2000; @Mullin2011] which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015] in North America.
 Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008].
@@
 All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009].
+In Quebec, several generations of seed orchards have already been established in white spruce.
+Thus, the current objectives of genetic selection for this species are to select trees with a higher wood quality  because white spruce has already reached a level considered interesting for traits such as rapid growth and volume of wood.
 \
 
-During the first half of the 20^th^ century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
-Therefore, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
-Stem vigor (biologic component) and wood quality (economic component) are two desirable traits that are taken into account in genetic selection.
-Biologically, the resistance to mortality by vigour is important to ensure to harvest leaving trees.
-Economically the value of the wood for end-use is important for industries.
+@mitchell1961 defined the quality of the wood as follows: "Wood quality is the resultant of physical and chemical characteristics possessed by a tree or a part of a tree that enable it to meet the property requirements for different end products".
+it is composed mainly of two components: economical and biological (stem vigor).
+These two components are threfore two desirable traits that are taken into account in genetic selection.
+Biologically, resistance to mortality by vigor is important to ensure the survival of the tree.
+Economically, the value of the wood for end-use is important for industries.
 Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
-More specifically the homogeneity of density and rings width within the stem has an high importance for stem vigor and wood quality because these two traits are related to it.
-Homogeneity refers here to a small interannual variation in wood density and rings width.\
-
-In view of the studies cited above, we understand that there might be a strong gain at selecting trees that are prone to produce wood with an homogeneous density to get a resistant and highly valuable forest in the next decades.
-Consequently, studies have been set up to understand the genetic sources of phenotype variations of wood quality [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+More specifically the low interannual variability of wood density and rings width has an high importance for stem vigor and wood quality because these two traits are related to it.
+In fact, the density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a low variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008].
+For example, @suarez2004 found that at the tree scale, individuals with variable growth were more prone to die of drought than more regularly growing trees.
+These studies have shown that the steady growth and therefore the low interannual variation wood density allows the tree to be more vigorous and therefore less prone to related mortality and environmental conditions.
+With regard to the economical context, interannual variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
+Indeed, low interannual variability of wood density affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bioenergy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
+For example, the low interannual variation in wood density is appreciated for sawnwood because sawyers place a high value on the low interannual variation wood density [Hassegawa et al (pending approval); @X].
+It is therefore important from both a biological and an end-use points of view to have a wood with a low interannual density variation within the stem.
+We understand that there might be a strong gain at selecting trees that are prone to produce wood with a low interannual variability of wood density to get a resistant and highly valuable forest in the next decades.
 For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Beaulieu2009; @Mullin2011].
-Hence, it is important to understand where do the variations of density come from in order to better integrated them into genetic selections.
-In white spruce, there is a typical low frequency radial pattern of variation from pith to bark [@Moore2011] with a density that decreases during the first years of growth and then an increasing from the beginning of the formation of mature wood [@panshin1980].
-These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
-This makes this trait attractive for genetic improvement.
-In addition to these low-frequency variations, there are high frequency inter-annual variations [@Nicault2010] that are influenced mainly by events to the tree such as the soil water availability and climatic conditions [@Nicault2010] that happen at high frequency.
-These variations are not under genetic control but under climate control meaning that climatic conditions also affect wood density.
-Some climatic factors such as drought can cause stress in trees and cause a variation in wood properties, including density, ring width and conductivity.
-Long-term growth analyses have shown that density may even increases with the temperature, the atmospheric CO~2~ content and the nitrogen deposition [@becker1995; @huang2007; @zaehle2014].
-Wood density can also be negatively correlated with growth.
-[@zhang1995].
-Therefore, with the aim of producing wood of homogeneous density it is essential to take climate factors into account in addition to genetic.\
-
- In fact, the density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a small variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008].
-For example, @suarez2004 found that at the tree scale, individuals with variable growth were more prone to die of drought than more regularly growing trees.
-These studies have shown that the steady growth and therefore the small variation in the density of the wood allows the tree to be more vigorous.
-With regard to the economical context, variation of density within a piece of wood may lead to deformation during transformation and inhomogeneous machining properties.
-Indeed, wood density homogeneity affects the performance of saw timber, the pulp yield and calorific values for biomass applications for bio energy [@Hernandez2001; @Macdonald2002; @Mckendry2002].
-For example, the low variation in wood density is appreciated for sawnwood because sawyers place a high value on the homogeneity of wood density [Hassegawa et al (pending approval); @X].
-It is therefore important from both a biological and an end-use points of view to have a wood with a uniform density within the stem.
+For wood quality, one of the characteristics that one wishes to select is the amplitude of the interannual variation of wood density, provided that it has sufficiently interesting heritability for a genetic selection.
 \
 
-A significant variation of the climate can cause a variation of the properties of the wood over several years.
-For example, several studies have shown that after the great drought of 2003 in Europe, the growth of forests has been affected over several years [@ciais2005; @granier2007; @reichstein2007].
-The ability of trees to tolerate these changes in a tree species or within a species is often related to the interaction between genotype and environment.
+Heritability is the probability that a trait of an individual is hereditarily transmitted by genetic factors.
+It is a statistical datum that serves to evaluate the part of genetic factors and that of environmental factors in the construction of the phenotype of a given population.
+Several studies have been set up to understand the genetic sources of phenotypic variations of wood quality traits and their heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+Thus, previous studies have shown that wood density is a trait under strong genetic control therefore strong heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
+It has been proven that the wood density is under genetic control, but we can ask ourselves the question of its amplitude of interannual variation.
+The interannual variation of wood density comes from two sources: the low-frequency radial profile and the high-frequency radial profile [@Nicault2010].
+The low-frequency radial profile is the radial variation of the density of pith-to-bark with a  high density during the first years of tree growth  decreases to a certain level and begins to increase again.
+In white spruce, this radial profile is the radial type II profile described by [@panshin1980].
+These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] because this functioning is internal to the tree and is transmitted to future generations.
+The high-frequency radial profile is the amplitude of the interannual variation of wood density.
+We did not find any studies on heritability of amplitude of interannual variation of wood density in white spruce.\
+
+Studies have linked annual environmental conditions such as soil water availability, climatic conditions that occur at a high frequency, and interannual variability wood density [@becker1995; @huang2007; @Nicault2010; @zaehle2014].
+Indeed, long-term growth analyzes have shown a link between wood density and local annual conditions such as summer temperature in white spruce [@darrigo1992] and other species [@becker1995; @huang2007; @zaehle2014].
+However, it seems to have some genetic control in the density sensitivity to environmental conditions.
 For example, studies have suggested a possible link between low sensitivity to drought and genotype [@ferrell1966; @teskey1987].
-This link is poorly controlled and therefore poorly known.
-It is in context that this study fits to try to advance the knowledge in this direction.\
+These studies show that although the interannual variation in density is under environmental control, its amplitude may be controlled by a genotype-environment interaction.
+Therefore, a question arises as to whether the amplitude of interannual variation is primarily under genetic control (heritability).
+Thus, in the context of climate change, it is crucial to know the level of heritability of the amplitude of interannual variation in wood density for the genetic selection of individuals with low amplitude of density variation.
+\
 
-In this study we quantified and compared heterogeneity of wood density from trees of known parents in order to figure out if some individuals produce more heterogeneous wood than others.
-Knowing the parents of each tree we tried to characterize the genetic contribution to the heterogeneity by comparing two growth sites we tried to characterize the environmental contribution to homogeneity.
-We ultimately aim to select and reproduce white spruce trees families of interest that are both low-sensitivity to climate change and genetically prone to make homogeneous wood.
-Thus, the most homogeneous families will be identified for reforestation and identify variables to consider in breeding of growing trees with less variable wood.
+In this study, we evaluated the amplitude of the interannual variation of wood density in white spruce and its share of heritability in the interannual conditions of the environment.
+We therefore compare two growth sites containing genotyped trees in order to characterize the genetic (heritability) and environmental contribution to the interannual variation of wood density.
+Thus, we hope to select and reproduce white spruce trees families of interest that are both low-sensitivity to climates changes and genetically prone to make low amplitude of interannual variation of wood density.
+If the amplitude of variation in wood density is under genetic control, tree families with the smallest variation amplitudes will be identified for reforestation.
 

### 2019-07-18 Article_AS.Rmd
Reorganisation intro

@@
 Thousands hectares of Canadian forest are harvested each year and an additional loss is due to insects, diseases and fires [@DRF2000; @NaturalResourcesCanada2019].
-It is important to increase tree growth to meet the growing demand of forest industry.
-The reforestation, when under human control, could ideally be made by using better trees than the former ones, i.e.
+In order to avoid deforestation, natural and artificial (planting or seeding) regeneration are used in Canada to compensate for harvests and forest losses.
+Artificial regeneration is used on more than half of the forests harvested in Canada [@RNC].
+One of the reasons for artificial regeneration is for better control of planted trees to meet the demands and requirements of forest industry.
+Indeed, artificial regeneration, could ideally be made by using better trees than the former ones, i.e.
 trees that growth faster, that are more resistant to perturbation and that are of better structural quality.
-To achieve this task, we must take into account the cambial activity that is governed by environmental conditions and genetics [@Dufour2010; @Rossi2012].
-Therefore, genetic selection is one of the available tools and is currently becoming a crucial and increasingly necessary for the development of the canadian forest [@Bonfils2004].
-In forestry, genetic selection is a process allows to select trees with characters of interest and to cross them or to reproduce them in order to obtain trees with interesting properties for the development of the forests or the quality of the wood for human use [@zobel1984; @DRF2000; @nanson2004].
+Thus, genetic selection is one of the tools for artificial regeneration and is becoming a crucial and increasingly necessary part of the development of canadian forest [@Bonfils2004].
+When used in forestry, genetic selection allows to select trees with characters of interest to cross or to reproduce them in order to obtain trees with desired properties for an improved development of the forests or an improved quality of the wood for human use [@zobel1984; @DRF2000; @nanson2004].
 \
@@
 During the first half of the 20^th^ century, most studies about genetic selection were dedicated on increasing wood volume production [@Dhir1976; @nienstaedt1985; @Li1997; @Koubaa2000] but with the increasing demand of high structural quality products and the fear of losing resource from natural causes, the aspects of wood quality became as important as wood volume [@Franceschini2012].
-Hence, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties.
+Hence, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties of commercial species such as white spruce.
 For example, improvement programs by genetic selection are in place for white spruce *(Picea glauca [Moench] Voss)* [@DRF2000; @Mullin2011] which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015] in North America.
-Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008].
-In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@nienstaedt1990].
-All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009].
 In Quebec, several generations of seed orchards have already been established in white spruce.
@@
 @mitchell1961 defined the quality of the wood as follows: "Wood quality is the resultant of physical and chemical characteristics possessed by a tree or a part of a tree that enable it to meet the property requirements for different end products".
-it is composed mainly of two components: economical and biological (stem vigor).
-These two components are threfore two desirable traits that are taken into account in genetic selection.
-Biologically, resistance to mortality by vigor is important to ensure the survival of the tree.
-Economically, the value of the wood for end-use is important for industries.
-Several traits were considered to assess wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
-More specifically the low interannual variability of wood density and rings width has an high importance for stem vigor and wood quality because these two traits are related to it.
-In fact, the density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a low variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008].
+We can consider that it is composed  of two components: one is biological and one economical.
+Biologically, resistance to mortality by vigor is important to ensure the survival of the tree and the value of the forest.
+Economically, the quality of the wood for end-use is important for industries as it impact the value of the end products.
+Several traits were considered to assess the different component of wood quality and the density is one of the most widely used indicators and arguably the most important one [@zobel1989; @Saranpaa1994; @Mckendry2002; @Macdonald2002; @Alteyrac2006].
+More specifically the low interannual variability of wood density and rings width has an high importance for stem vigor and wood quality.
+First of all, it is important to understand that, the interannual variation of wood density comes from two sources: the low-frequency radial profile and the high-frequency radial profile [@Nicault2010].
+The low-frequency radial profile is the radial variation of the density of pith-to-bark with a high density during the first years of tree growth  decreases to a certain level and begins to increase again.
+In white spruce, this radial profile is the radial type II profile described by @panshin1980.
+These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] because this functioning is internal to the tree and is transmitted to future generations.
+The high-frequency radial profile is the amplitude of the interannual variation of wood density i.e.
+the density difference between two consecutive growing years (fig 1).
+
+![Schematic curves showing various signals in a dendrochronological series: low frequency age signal and high frequency climate signal.
+(modified from @Nicault2010)](graph_table/signal.png){width=200px; height=300px}
+
+This high frequency signal is related to climatic conditions [@Nicault2010].
+Therefore, significant inter-annual fluctuations in climate lead to an increase in inter-annual variation in wood density.
+In the coming years, the range of year-to-year variation in wood density may change due to climate changes.
+Indeed, wood density is correlated with the growth of the tree [@corriveau1991] and it has been observed that trees with a low variation in growth and therefore in density are also the most resistant to mortality [@ogle2000; @suarez2004; @hartmann2008].
 For example, @suarez2004 found that at the tree scale, individuals with variable growth were more prone to die of drought than more regularly growing trees.
@@
 It is therefore important from both a biological and an end-use points of view to have a wood with a low interannual density variation within the stem.
-We understand that there might be a strong gain at selecting trees that are prone to produce wood with a low interannual variability of wood density to get a resistant and highly valuable forest in the next decades.
+Therefore, we understand that there might be a strong gain at selecting trees that are prone to produce wood with a low interannual variability of wood density to get a resistant and highly valuable forest in the next decades.
 For example, future genetic selections in breeding programs for different species of spruce trees in Quebec will consider wood properties such as wood density [@Beaulieu2009; @Mullin2011].
-For wood quality, one of the characteristics that one wishes to select is the amplitude of the interannual variation of wood density, provided that it has sufficiently interesting heritability for a genetic selection.
+For white spruce wood quality, one of the characteristics that one wishes to select is the amplitude of the interannual variation of wood density, provided that it has sufficiently interesting heritability for a genetic selection.
 \
 
-Heritability is the probability that a trait of an individual is hereditarily transmitted by genetic factors.
-It is a statistical datum that serves to evaluate the part of genetic factors and that of environmental factors in the construction of the phenotype of a given population.
-Several studies have been set up to understand the genetic sources of phenotypic variations of wood quality traits and their heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
-Thus, previous studies have shown that wood density is a trait under strong genetic control therefore strong heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
-It has been proven that the wood density is under genetic control, but we can ask ourselves the question of its amplitude of interannual variation.
-The interannual variation of wood density comes from two sources: the low-frequency radial profile and the high-frequency radial profile [@Nicault2010].
-The low-frequency radial profile is the radial variation of the density of pith-to-bark with a  high density during the first years of tree growth  decreases to a certain level and begins to increase again.
-In white spruce, this radial profile is the radial type II profile described by [@panshin1980].
-These variations are under genetic control and are consequently moderately to highly heritable [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006] because this functioning is internal to the tree and is transmitted to future generations.
-The high-frequency radial profile is the amplitude of the interannual variation of wood density.
-We did not find any studies on heritability of amplitude of interannual variation of wood density in white spruce.\
+Heritability is the probability that a trait of an individual is hereditarily transmitted by genes.
+It is a statistical information that serves to evaluate the part of genome and that of environmental conditions in the construction of the phenotype of a given population.
+Moreover, several studies have been focused on the understanding the genetic sources of phenotypic variations of wood quality traits (eg.
+density) and their heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+Previous studies have shown that wood density is a trait under strong genetic control and therefore have a strong heritability [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006].
+As mentioned above, the low-frequency radial variation is age-related and under genetic control with strong heritability [@Corriveau1971; @Li1993; @Hernandez2001; @Beaulieu2003; @Lenz2013; @McLean2016].
+However, we did not find any studies on heritability of amplitude of interannual variation of wood density in white spruce.\
 
-Studies have linked annual environmental conditions such as soil water availability, climatic conditions that occur at a high frequency, and interannual variability wood density [@becker1995; @huang2007; @Nicault2010; @zaehle2014].
+Studies have linked annual environmental conditions such as soil water availability, climatic conditions and interannual variability wood density (high frequency signal) [@becker1995; @huang2007; @Nicault2010; @zaehle2014].
 Indeed, long-term growth analyzes have shown a link between wood density and local annual conditions such as summer temperature in white spruce [@darrigo1992] and other species [@becker1995; @huang2007; @zaehle2014].
-However, it seems to have some genetic control in the density sensitivity to environmental conditions.
+However, the results of several studies suggest a genetic control of density sensitivity to environmental conditions.
 For example, studies have suggested a possible link between low sensitivity to drought and genotype [@ferrell1966; @teskey1987].
-These studies show that although the interannual variation in density is under environmental control, its amplitude may be controlled by a genotype-environment interaction.
-Therefore, a question arises as to whether the amplitude of interannual variation is primarily under genetic control (heritability).
-Thus, in the context of climate change, it is crucial to know the level of heritability of the amplitude of interannual variation in wood density for the genetic selection of individuals with low amplitude of density variation.
+These studies show that although the interannual variation in density (high frequency signal) is under environmental control, its amplitude may be controlled by a genotype-environment interaction.
+Therefore, a question arises as to whether the amplitude of interannual variation is primarily under genetic control (heritability) or environmental.
+Knowing that the interannual variation in density is controlled by the annual environmental conditions, in the context of climatic changes that are likely to increase the interannual variation of environmental conditions, it is crucial to know whether the amplitude of variation in wood density is under genetic control.
 \
 
-In this study, we evaluated the amplitude of the interannual variation of wood density in white spruce and its share of heritability in the interannual conditions of the environment.
-We therefore compare two growth sites containing genotyped trees in order to characterize the genetic (heritability) and environmental contribution to the interannual variation of wood density.
+In this study, we evaluated the amplitude of the interannual variation (high frequency signal) of wood density in white spruce and its share of heritability in the interannual conditions of the environment.
+We therefore compare two growth sites containing genotyped trees in order to characterize the genetic (heritability) and environmental contribution to the amplitude of the interannual variation of wood density.
 Thus, we hope to select and reproduce white spruce trees families of interest that are both low-sensitivity to climates changes and genetically prone to make low amplitude of interannual variation of wood density.
-If the amplitude of variation in wood density is under genetic control, tree families with the smallest variation amplitudes will be identified for reforestation.
+If the range of variation in wood density is under genetic control, tree families with the smallest variation ranges can be identified for reforestation.
 
@@
 
-The material used was sampled from an experimental provenance-progeny test plantation designed to assess the genetic merit of first generation selections, which had been subdivided into breeding groups and delineated by provenance or geographic region.
+The material used was sampled from an experimental provenance-progeny test plantations designed to assess the genetic merit of first generation selections, which had been subdivided into breeding groups and delineated by provenance or geographic region.
 The details of the setting up of the experiment are described in @beaulieu1996.
@@
 Each parent tree has been used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups.
-Genetic testing was conducted in 1999 on 2-year-old nursery seedlings near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1).
+Genetic testing were set up in 1999 on 2-year-old seedlings from these experimental plantations near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1).
 Trees were assigned to plots in rows of five trees / plot (spacing of 2 m × 2 m) [@beaulieu2014].
-In this study, 93 full-brother families for a total of 2365 trees.
+In this study, we were used 93 full-brother families for a total of 2365 white spruce trees.
 The sampled trees were drawn from 18 blocks over the two sites and 9–14 individuals were sampled per full-sib cross.
@@
 
-|Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |
-|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|
-|Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |
-|Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |
+|Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |Ecological region      |
+|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|:----------------------|
+|Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |balsam fir-yellow birch|
+|Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |maple–basswood         |
 Table: Location and climatic conditions* (over 1981–2010) of the two plantation sites.
@@
 
-Wood cores were extracted à une épaisseur de 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground).
+Wood cores were extracted to a thickness of 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground).
 Tree cores are extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself.
 These cores were used to determine ring width and wood density.
-Density was determined by scanning incremental cores from bark to bark along the radial face at a resolution of 25 $\mu$m in the QTRS-01X ring analyzer (Quintek Measurement Systems Inc., Knoxville, TN, USA).
+Density was determined by scanning incremental cores from pith-to-bark along the radial face at a resolution of 25 $\mu$m in the QTRS-01X ring analyzer (Quintek Measurement Systems Inc., Knoxville, TN, USA).
 A weighted mean density line was obtained by summing the product between the annual density measurements and the ratio of the respective radial areas to the total radial area.
@@
 From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines.
-Ring width determined through information of densitometry profiles.\
+Ring width determined through information of densitometry profiles.
+The transition between earlywood and the latewood is determined using the "Unix" script.
+This script used a transition density threshold ($D_t$) determined for each ring as follows:
+
+$$D_t= \frac{(D_{min}+D_{max})}{2}$$
+Where $D_t$ is ring density threshold, $D_{min}$ is ring minimum density and $D_{max}$ is ring maximum density. Latewood started when these conditions were met:
+
+1. $D_x < D_t$
+
+2. $D_{x+1} >= D_t$
+
+3. $D_{x+2} >= D_t$
+
+Where $D_x$ is current reading and $D_t$ is next reading. \
 
@@
 Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018] (Eq.3).
-Previous studies have used this method to describe the density profile of different spruce species that have a similar profile[@franceschini2010; @gardiner2011].
-However, when modeling density profiles we found that the variation between individual discs follows different patterns from the marrow to the bark between tree.
+Previous studies have used this method to describe the density profile of different spruce species that have a similar profile [@franceschini2010; @gardiner2011].
+However, when modeling density profiles we found that the variation between individual discs follows different patterns from the pith-to-bark between tree.
 We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
@@
 We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
-The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999].
+The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values in addition to the explanatory variables, fit indices (R^2^) [@Pape1999].
 As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores.
@@
 Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.
-The graphic description in figure 2 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).
+The graph in figure 2 describes the three parameters of our model and allows us to understand how each parameter influences our model (Fig 2).
 
@@
 
-## Determination of the wood density variability index
+## Determination of interannual variation amplitude of wood density index
 After subtraction of the typical density, it is observed that the residues remain variable.
 Knowing that the low frequency signal is subtracted, this variation can only be due to the high frequency signal which is the interannual variation.
-As explained earlier in the introduction, low interannual variability is an important criterion for wood quality.
-In order to identify tree families with less interannual variability in density, we have created a viability index.
-The variability index is an index to determine the uniformity in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
-The residues of our model were used to obtain the wood density variability index, which was determined using an equation.
+As explained earlier in the introduction, low interannual variation is an important criterion for wood quality.
+In order to identify tree families with low interannual variation amplitude in density, we have created a viability index.
+The interannual variation amplitude of wood density index is an index to determine the variability in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
+The residues of our model were used to obtain the interannual variation amplitude of wood density index, which was determined using an equation.
 We subtracted the density residues values of year $n-1$ from year $n$.
@@
 
-Where V is the variability index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
+Where V is the interannual variation amplitude of wood density index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
 
@@
 We therefore used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites.
-We calculated this coefficient for the variability index and the different parameters of the model.
-The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites and the three parameters to estimate from the model.
+We calculated this coefficient for the interannual variation amplitude of wood density index and the different parameters of the model.
+The elements for which we calculated Spearman’s rho are: the interannual variation amplitude of wood density index for each of the tree families between two study sites and the three parameters to estimate from the model.
 The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho.
@@
 
- $$h_2= 2 \frac{\sigma^2_f}{\sigma^2_T}$$
+ $$h_2= 2 \frac{\sigma^2_f}{\sigma^2_t}$$
 
-Where $h_2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_T$ is total variance.
+Where $h_2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_t$ is total variance.
 ASReml3 package [@Butler2018] in the R statistical programming environment [@R2018] was used to estimate trait genetic parameters based on the residual maximum likelihood.
@@
 
-|Site                     |Parameter |    Min|   Mean|    Max|
+|Site                     |Parameter |Min    |Mean   |Max    |
 |:------------------------|:---------|------:|------:|------:|
@@
 ![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px}.\
+Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px} \
 
 ![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px}.
-\
+Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px} \
 
@@
 
-|Site                     |  Min|  Mean|    Max|
+|Site                     |Min  |Mean  |Max    |
 |:------------------------|----:|-----:|------:|
@@
 
-|Rank correlation  |  Rho| P-value|
+|Rank correlation  |Rho  |P-value |
 |:-----------------|----:|-------:|
@@
 |Trait             |Combined sites |Asselin |Saint-Casimir |
-|:----------------|:--------------|:-------|:-------------|
-|Wood density     |0.55           | 0.58   |0.70          |
+|:-----------------|:--------------|:-------|:-------------|
+|Wood density      |0.55           |0.58    |0.70          |
 |Width             |0.23           |0.32    |0.43          |

### 2019-08-09 Article_AS.Rmd
Mat et met

@@
 Hence, conifer breeding programs are more and more considering wood quality as selection criteria to improve end-use properties of commercial species such as white spruce.
-For example, improvement programs by genetic selection are in place for white spruce *(Picea glauca [Moench] Voss)* [@DRF2000; @Mullin2011] which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015] in North America.
+For example, improvement programs by genetic selection are in place for white spruce (*Picea glauca* [Moench] Voss) [@DRF2000; @Mullin2011] which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015] in North America.
 In Quebec, several generations of seed orchards have already been established in white spruce.
@@
 In this study, we were used 93 full-brother families for a total of 2365 white spruce trees.
-The sampled trees were drawn from 18 blocks over the two sites and 9–14 individuals were sampled per full-sib cross.
+The sampled trees were drawn from 18 plots over the two sites and 9–14 individuals were sampled per full-sib cross.
 Trees used in this study are exclusively full-brother families.\
@@
 Wood cores were conditioned to about 7% moisture before scanning.
-Thus, the determination of the density of the wood makes it possible to calculate a quality index of it.
+Thus, the determination of the wood density makes it possible to calculate a quality index of it.
 From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines.
@@
 As part of this study, it is essential to accurately identify the year of formation of the annual rings measured.
-To this end, we used the technique of Cross-dating which consists in obtaining the concordance of the number of rings and of their width pattern to avoid counting errors or those caused by missing rings or by false annual rings.
-Crossdating is the most basic principle of dendrochronology.
-Cross-dating is a technique that ensures each individual tree ring is assigned its exact year of formation.
-This is accomplished by matching patterns of wide and narrow rings between cores from the same tree, and between trees from different locations.
-To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983].
+To this end, we used the technique of Crossdating which consists in obtaining the concordance of the number of rings and of their width pattern to avoid counting errors or those caused by missing rings or by false annual rings.
+Crossdating  is one of the most important basic principles of dendrochronology.
+This technique ensures that each tree ring is assigned to the exact year of its formation.
+This is accomplished by matching patterns of wide and narrow rings between cores from the same tree, and between trees from different positions.
+To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for crossdating and quality control of annual ring measurements [@HOLMES1983].
 Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be corrected.
@@
 A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations.
-Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018] (Eq.3).
+Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
 Previous studies have used this method to describe the density profile of different spruce species that have a similar profile [@franceschini2010; @gardiner2011].
-However, when modeling density profiles we found that the variation between individual discs follows different patterns from the pith-to-bark between tree.
+However, when modeling density profiles we found that the variation between individual cores follows different patterns from the pith-to-bark between trees.
 We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
-To work around this problem, we chose to model each of the profiles separately.
-We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core .
-The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values in addition to the explanatory variables, fit indices (R^2^) [@Pape1999].
+To circumvent this problem, we chose to model each of the profiles separately.
+We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core.
+Model fit of the model was evaluated by visual analysis of the model residual curves against adjusted values in addition to the explanatory variables, fit indices (R^2^) [@Pape1999].
 As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores.
@@
 where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term.
-Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.
+Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0; $b_1$ denotes the scale parameter of the early decrease in RD.
+$b_2$ denotes the rise in RD at the end of the juvenile wood and at the beginning of the mature wood formation.
+The error term $e$ was assumed to be normally distributed, with N (0, $\sigma^{2}$).
+The first exponential term $b_0 \times \exp\left(\frac{b_1}{CA}\right)$ has a value which decreases rapidly in the first few annual rings and then gradually approaches an asymptote close to zero.
 The graph in figure 2 describes the three parameters of our model and allows us to understand how each parameter influences our model (Fig 2).
@@
 
-## Determination of interannual variation amplitude of wood density index
+## Determination of interannual radial variation amplitude of wood density index
 After subtraction of the typical density, it is observed that the residues remain variable.
-Knowing that the low frequency signal is subtracted, this variation can only be due to the high frequency signal which is the interannual variation.
-As explained earlier in the introduction, low interannual variation is an important criterion for wood quality.
-In order to identify tree families with low interannual variation amplitude in density, we have created a viability index.
+Knowing that the low frequency signal is subtracted, this radial variation can only be due to the high frequency signal which is the interannual radial variation.
+As explained earlier in the introduction, low interannual radial variation is an important criterion for wood quality.
+In order to identify tree families with low interannual radial variation amplitude in density, we have created a interannual radial variation amplitude index.
 The interannual variation amplitude of wood density index is an index to determine the variability in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
@@
 
-Where V is the interannual variation amplitude of wood density index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
+Where V is the interannual radial variation amplitude of wood density index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
 
@@
 ## Quantitative genetic analyses
-Variance and covariance for genetics were estimated using mixed models [@Lenz2013] of the R software. The equation used is as follows:
+Variance and covariance for genetics were estimated using linear mixed-effects model [@Lenz2013] of the R software. The equation used is as follows:
 
-$$ y_{ijkl} = \mu + s_i + b_{(i)j} + f_k + bf_{(i)jk} + e_{ijkl}$$
+$$ Y_{ijkl} = \mu + s_i + b_{(i)j} + f_k + bf_{(i)jk} + e_{ijkl}$$
 
-where $y_{ijkl}$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
+where $Y_{ijkl}$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
 In a full-sib design with random mating, members of the same family are presumed to share half of their genes [@white2007]. Therefore, individual heritability was obtained as follows:
@@
 The model we have developed has made it possible to optimize the density of the wood for each core.
-In table 1, we observe that the average value of the parameters differs between Asselin and Saint-Casimir.
-For the parameters $b_0$ and $b_1$, the average is higher in Saint-Casimir and for parameter $b_2$ it is higher in Asselin.\
+A summary of the parameter values obtained of the model fitting process (Eq.
+1) can be found in Table 2.
+We observe that the average value of the model parameters differs between Asselin and Saint-Casimir.
+For the parameters $b_0$ and $b_1$, the average is higher in Saint-Casimir and for parameter $b_2$ it is higher in Asselin.
+Although there are small differences in the parameters between the two sites, these differences are very small.
+The fit of the model is therefore similar at both sites.\
 
@@
 ![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px} \
+Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px}
 
 ![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px} \
+Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px}
 
 When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings.
-It goes from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to kg/m^3^ for the Asselin site.
-Then, this decrease was followed by a slight increase until the twelfth year.
+It declined from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to 375 kg/m^3^ for the Asselin site.
+Then, this decrease was followed by a slow increase until the twelfth year.
 \
 
-Model predictions are similar to data trends across sites (fig 3&4).
+Model predictions are similar to data trends across sites (figs 4 and 5).
 A decrease in density is observed in the first years of growth and then an increase in density from a certain cambial age.
-Whether it is the data or the predictions of the model, we observe a difference between the Asselin (fig 3) and Saint-Casimir (fig 4) sites.
+Whether it is the data or the predictions of the model, we observe a difference between the Asselin (fig 4) and Saint-Casimir (fig 5) sites.
 The increase in density after the decrease of the first years of growth is greater in Saint-Casimir than in Asselin.
@@
 |:------------------------|----:|-----:|------:|
-|Asselin (Témiscouata)    |  0.0| 417.5| 1155.8|
+|Asselin (Témiscouata)    |116.5| 417.9| 1155.8|
 |Saint-Casimir (Portneuf) | 54.9| 339.8|  746.0|
-|Combined                 |  0.0| 375.9| 1155.8|
+|Combined                 | 54.9| 375.9| 1155.8|
 Table: Summary table of variability index
@@
 When comparing the index of variability between study sites, there is greater variability in Asselin than in Saint-Casimir.
-The maximum value of 1155.8 in Asselin, for example, is 746, which represents a huge difference.
+A summary of the density interannual amplitude index values can be found in table 3.
+The maximum value is 1155.8 in Asselin, and 746 for the Saint-Casimir site, which represents a huge difference.
+\
 
@@
 
-The results of the rank correlation coefficient are summarized in Table 3.
+The results of the rank correlation coefficient are summarized in Table 4.
 These results show an absence of correlation for the variability index (p-value = 0.43) and a strong correlation for all the parameters of the model (p-value = 2.2 10^-16^).
 The value of Rho is very similar between the three parameters.
+\
 
@@
 The uniformity index has almost zero heritability regardless of the study site.\
+
+\clearpage
 

### 2019-08-27 Article_AS.Rmd
graphs

@@
 Crosses were made according to a partial diallel mating design within each of the six different breeding groups or sublines [@Burdon1983], which were designed to limit future inbreeding to within group.
-Each parent tree has been used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups.
-Genetic testing were set up in 1999 on 2-year-old seedlings from these experimental plantations near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1).
+Each parent tree was used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups.
+Genetic testing were set up in 1999 on 2-year-old seedlings from these experimental plantations near Asselin balsam fir-yellow birch ecological domain, and Saint-Casimir located in the maple–basswood ecological domain, Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1).
 Trees were assigned to plots in rows of five trees / plot (spacing of 2 m × 2 m) [@beaulieu2014].
@@
 |Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |Ecological region      |
-|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|:----------------------|
+|:-------------|-----------:|--------:|--------:|-------:|--------:|--------:|----------------:|:----------------------|
 |Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |balsam fir-yellow birch|
 |Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |maple–basswood         |
-Table: Location and climatic conditions* (over 1981–2010) of the two plantation sites.
+Table: Location and climatic conditions (over 1981–2010) of the two plantation sites. LAT, latitude; LON, longitude; ELV, elevation; MAT, mean annual temperature; TAP, total annual precipitation.
 
-*LAT, latitude; LON, longitude; ELV, elevation; MAT, mean annual temperature; TAP, total annual precipitation.
+## Sample preparation and scanning
 
 Wood cores were extracted to a thickness of 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground).
-Tree cores are extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself.
+Wood cores were extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself.
 These cores were used to determine ring width and wood density.
@@
 Wood cores were conditioned to about 7% moisture before scanning.
-Thus, the determination of the wood density makes it possible to calculate a quality index of it.
-From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines.
+Determination of the wood density makes it possible to calculate a quality index of it.
+From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring.
 Ring width determined through information of densitometry profiles.
-The transition between earlywood and the latewood is determined using the "Unix" script.
-This script used a transition density threshold ($D_t$) determined for each ring as follows:
+The transition between earlywood and the latewood is determined using a script developed, to check for possible errors (missing rings, abnormally low or high readings) and calculate the transition point between the initial wood and the final wood.
+The script uses the average between the minimum density and the maximum density to obtain the transition density threshold ($D_t$).
 
-$$D_t= \frac{(D_{min}+D_{max})}{2}$$
-Where $D_t$ is ring density threshold, $D_{min}$ is ring minimum density and $D_{max}$ is ring maximum density. Latewood started when these conditions were met:
+<!--- $$D_t= \frac{(D_{min}+D_{max})}{2}$$
+Where $D_t$ is ring density threshold, $D_{min}$ is ring minimum density and $D_{max}$ is ring maximum density. Latewood started when this condition is met:
 
-1. $D_x < D_t$
+$$D_i < D_t \&\& D_{i+1} >= D_t \&\& D_{i+2} >= D_t$$
 
-2. $D_{x+1} >= D_t$
-
-3. $D_{x+2} >= D_t$
-
-Where $D_x$ is current reading and $D_t$ is next reading. \
+Where $D_i$ is the ith density data. ---->
 
@@
 To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for crossdating and quality control of annual ring measurements [@HOLMES1983].
-Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be corrected.
+Annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be corrected.
 Any incomplete or false rings were excluded from the dataset.
@@
 A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations.
-Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018].
-Previous studies have used this method to describe the density profile of different spruce species that have a similar profile [@franceschini2010; @gardiner2011].
-However, when modeling density profiles we found that the variation between individual cores follows different patterns from the pith-to-bark between trees.
-We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others.
-To circumvent this problem, we chose to model each of the profiles separately.
+We had hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks.
+Unlike some previous studies [@franceschini2010; @gardiner2011] that used mixed models to describe the radial density profile, we adjusted the curve of our equation to each core for a better description of the profiles.
 We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core.
-Model fit of the model was evaluated by visual analysis of the model residual curves against adjusted values in addition to the explanatory variables, fit indices (R^2^) [@Pape1999].
 As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores.
@@
 
-$$RD = b_0 \times \exp\left(\frac{b_1}{CA}\right) + \frac{CA}{b_2} + e $$
+$$RD(t) = b_0 \times \exp\left(\frac{b_1}{t}\right) + \frac{t}{b_2} + e $$
 
-where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term.
-Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0; $b_1$ denotes the scale parameter of the early decrease in RD.
+where RD is the mean ring density (kg.m^-3^), t is cambial age (years), $b_0$, $b_1$ and $b_2$ are adjusted parameters, and $e$ is the residual error term.
+Parameter $b_0$ denotes the intercept when t approaches a theoretical value of 0; $b_1$ denotes the scale parameter of the early decrease in RD.
 $b_2$ denotes the rise in RD at the end of the juvenile wood and at the beginning of the mature wood formation.
@@
 ## Determination of interannual radial variation amplitude of wood density index
-After subtraction of the typical density, it is observed that the residues remain variable.
-Knowing that the low frequency signal is subtracted, this radial variation can only be due to the high frequency signal which is the interannual radial variation.
-As explained earlier in the introduction, low interannual radial variation is an important criterion for wood quality.
-In order to identify tree families with low interannual radial variation amplitude in density, we have created a interannual radial variation amplitude index.
-The interannual variation amplitude of wood density index is an index to determine the variability in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions.
-The residues of our model were used to obtain the interannual variation amplitude of wood density index, which was determined using an equation.
+Type II variation profile (low frequency signal) being subtracted from the residues represents the high frequency signal which is the interannual variation.
+In order to identify tree families with low interannual variation amplitude in density, we have created a index.
+This index makes it possible to determine the amplitude of the interannual variation of the density in order to identify the families of trees least sensitive to the variations of the environmental conditions.
 We subtracted the density residues values of year $n-1$ from year $n$.
@@
 
-Where V is the interannual radial variation amplitude of wood density index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
+Where V is the index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$.
 
 ## Spearman’s rank correlation coefficient (Spearman’s rho) 
-Spearman’s rank correlation coefficient is a correlation coefficient to describe the link between two variables by means of a monotonic function.
-We therefore used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites.
-We calculated this coefficient for the interannual variation amplitude of wood density index and the different parameters of the model.
-The elements for which we calculated Spearman’s rho are: the interannual variation amplitude of wood density index for each of the tree families between two study sites and the three parameters to estimate from the model.
+We used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites.
+The elements for which we calculated Spearman’s rho are: V for each of the tree families between two study sites and the three parameters to estimate from the model.
 The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho.
@@
 
-$$rho = \frac{cov(rg_X,rg_Y)}{\sigma_{rg_X} \times\sigma_{rg_Y}}$$
+<!---$$rho = \frac{cov(rg_X,rg_Y)}{\sigma_{rg_X} \times\sigma_{rg_Y}}$$
 Where $rho$ is a nonparametric measure of rank correlation, $cov(rg_X,rg_Y)$ is the covariance of the rank variables.
 $\sigma_{rg_X}$ and $\sigma_{rg_Y}$ are the standard deviations of the rank variables.
+----->
 
 ## Quantitative genetic analyses
-Variance and covariance for genetics were estimated using linear mixed-effects model [@Lenz2013] of the R software. The equation used is as follows:
+Variance components for genetic analysis were estimated using ASReml 3 [@Butler2018] based on linear mixed-effects model using as follows:
 
@@
 
-where $Y_{ijkl}$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual error.
-In a full-sib design with random mating, members of the same family are presumed to share half of their genes [@white2007]. Therefore, individual heritability was obtained as follows:
+where $Y_{ijkl}$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual effect.
+In a full-sib design with random mating, members of the same family are presumed to share half of their genes [@white2007].
+Therefore, Individual-tree narrow-sense heritability ($hi^2$), obtained based on each fitted regression function, was calculated using the following equation:
 
- $$h_2= 2 \frac{\sigma^2_f}{\sigma^2_t}$$
+ $$hi^2= 2 \frac{\sigma^2_f}{\sigma^2_t}$$
 
-Where $h_2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_t$ is total variance.
-ASReml3 package [@Butler2018] in the R statistical programming environment [@R2018] was used to estimate trait genetic parameters based on the residual maximum likelihood.
+Where $hi^2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_t$ is total variance.
 
@@
 
-The model we have developed has made it possible to optimize the density of the wood for each core.
+The model we have developed has made it possible to adjust wood density for each core.
 A summary of the parameter values obtained of the model fitting process (Eq.
-1) can be found in Table 2.
+1) can be found in figure 4&5.
 We observe that the average value of the model parameters differs between Asselin and Saint-Casimir.
@@
 Although there are small differences in the parameters between the two sites, these differences are very small.
-The fit of the model is therefore similar at both sites.\
+The fit of the model is therefore similar at both sites.
 
-|Site                     |Parameter |Min    |Mean   |Max    |
-|:------------------------|:---------|------:|------:|------:|
-| Asselin (Témiscouata)   |$b_0$     | 257.26| 302.78| 369.78|
-|                         |$b_1$     |  -0.93|   0.97|   5.36|
-|                         |$b_2$     | -81.83|  17.02|  61.34|
-| Saint-Casimir (Portneuf)|$b_0$     | 264.88| 304.40| 353.90|
-|                         |$b_1$     |  -1.34|   1.09|   7.04|
-|                         |$b_2$     | -50.87|  14.87|  61.34|
-Table: Summary table of parameters
+![Summary graphic of parameters in Asselin.](graph_table/parameter.asselin.png){width=335px; height=330px}
+
+![Summary graphic of parameters in Saint-Casimir.](graph_table/parameters.casimir.png){width=335px; height=330px}
 
 ![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px}
+Fine lines show densities for individual stands of the samples for the Asselin site.
+The bold lines shows the density predicted by Model.](graph_table/Density_asselin.png){width=330px; height=330px}
 
 ![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px}
+Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model.](graph_table/Density_casimir.png){width=330px; height=330px}
 
@@
 
-Model predictions are similar to data trends across sites (figs 4 and 5).
+Model predictions are similar to data trends across sites (figs 6 and 7).
 A decrease in density is observed in the first years of growth and then an increase in density from a certain cambial age.
-Whether it is the data or the predictions of the model, we observe a difference between the Asselin (fig 4) and Saint-Casimir (fig 5) sites.
+Whether it is the data or the predictions of the model, we observe a difference between the Asselin (fig 6) and Saint-Casimir (fig 7) sites.
 The increase in density after the decrease of the first years of growth is greater in Saint-Casimir than in Asselin.
@@
 
-|Site                     |Min  |Mean  |Max    |
-|:------------------------|----:|-----:|------:|
-|Asselin (Témiscouata)    |116.5| 417.9| 1155.8|
-|Saint-Casimir (Portneuf) | 54.9| 339.8|  746.0|
-|Combined                 | 54.9| 375.9| 1155.8|
-Table: Summary table of variability index
+![Summary graphic of index V per site.](graph_table/indexV.png){width=335px; height=330px}
 
-When comparing the index of variability between study sites, there is greater variability in Asselin than in Saint-Casimir.
-A summary of the density interannual amplitude index values can be found in table 3.
+When comparing the index V between study sites, there is greater variability in Asselin than in Saint-Casimir.
+A summary of index V values can be found in figure 8.
 The maximum value is 1155.8 in Asselin, and 746 for the Saint-Casimir site, which represents a huge difference.
@@
 |Rank correlation  |Rho  |P-value |
-|:-----------------|----:|-------:|
+|:-----------------|----:|:-------|
 |Density residus   | 0.57|8.7^-09^|
-|Variability index | 0.08|    0.43|
+|Variability index | 0.08|0.43    |
 |$b_0$             | 0.70|2.2^-16^|
@@
 
-The results of the rank correlation coefficient are summarized in Table 4.
-These results show an absence of correlation for the variability index (p-value = 0.43) and a strong correlation for all the parameters of the model (p-value = 2.2 10^-16^).
+The results of the rank correlation coefficient are summarized in Table 2.
+These results show an absence of correlation for the index V (p-value = 0.43) and a strong correlation for all the parameters of the model (p-value = 2.2 10^-16^).
 The value of Rho is very similar between the three parameters.
@@
 
-|Trait             |Combined sites |Asselin |Saint-Casimir |
-|:-----------------|:--------------|:-------|:-------------|
-|Wood density      |0.55           |0.58    |0.70          |
-|Width             |0.23           |0.32    |0.43          |
-|$b_0$             |0.20           |0.20    |0.3           |
-|$b_1$             |0.06           |0.10    |0.06          |
-|$b_2$             |0.04           |0.10    |0.03          |
-|Variability index |1.10^-6^       |6.10^-6^|3.10^-6^      |
-Table: Abstract of the heritability of the density, the ring width, parameters $b_0$, $b_1$, $b_2$ and the Variability index.
+![Abstract of the heritability ($hi^2$) of the density, the ring width, parameters $b_0$, $b_1$, $b_2$, index V and standard error per site.](graph_table/heritability.png){width=335px; height=330px}
 
-Heritability is higher in each site separately than that of the combined sites for all traits.
+Heritability is higher in each site separately than that of the combined sites for all traits (fig 9).
 Wood density is the only character with strong heritability for each of the sites as well as for the combined sites.

### 2019-09-11 Article_AS.Rmd
modif discussion

@@
 |Density residus   | 0.57|8.7^-09^|
-|Variability index | 0.08|0.43    |
+|Index V           | 0.08|0.43    |
 |$b_0$             | 0.70|2.2^-16^|
@@
 When parameters $b_1$ and $b_2$, their heritability is very low.
-The uniformity index has almost zero heritability regardless of the study site.\
+The index V has almost zero heritability regardless of the study site.\
 
@@
 
-Like the density profiles, variability index is different between the two sites, despite the fact that they are the same families of trees.
-Since the variability index is calculated from the model residuals, the difference can be explained by the fact that the variability that is related to the internal control of the trees has been subtracted from the model and that the remaining one is more controlled by external phenomena related to the study sites.
-Variability index being more under environmental control than genetic, this explain the lack of heritability of this index.
-It is not surprising that the variability index is under environmental control because density is mainly controlled by genetics but partly also by growth that is mainly controlled by environmental conditions.
-The model has made it possible to subtract the typical pattern of the density variation which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control.
+Like the density profiles, index V is different between the two sites, despite the fact that they are the same families of trees.
+Since the index V is calculated from the model residuals, the difference can be explained by the fact that the variability that is related to the internal control of the trees has been subtracted from the model and that the remaining one is more controlled by external phenomena related to the study sites.
+Index V being more under environmental control than genetic, this explain the lack of heritability of this index.
+The model has made it possible to subtract the type II pith-to-bark profile of the density variation which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control.
 Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.\
 
-Variability index being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index.
-This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the variability index is under environmental control.
-The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites taking into account the extreme variability of the different density profiles.\
+Index V being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index.
+This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the index V is likely under environmental control.
+The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites.
+Spearman's Rank Correlation Coefficient highlights the interaction between family and environment.
+The interaction is more or less important from one factor to another probably because some factors are more controlled by the environment and others by genetics.\
 
+Genetic parameter estimates were used to evaluate the variation of density as a selection trait for wood quality.
 With regard to heritability of density, it is high at both sites.
@@
 All of these studies reported values between 0.36 and 0.80.
+Average density of our data is therefore highly controlled by genetics as might be expected.
 In contrast to density, annual ring width has low to moderate heritability in each site and also for combined sites.
@@
 This result suggests that, if trees are to be selected for reforestation as a result of this study, the territory be divided into relatively homogeneous subunits, within which the recommended sources will be able to keep their superiority and can be used without risk of maladjustment [@beaulieu1989stabilite].
-
 The parameters of the model, as for them, they have a low heritability either for each site separately or for the combined sites.
 The parameter $b_0$, has a higher heritability than the other parameters but remains low.
-This can be explained by the fact that this parameter is very little influenced by cambial age.
+This can be explained by the fact that this parameter is very little influenced by cambial age because it is described the density close to pith.
 In fact, $b_0$ is the ordinate at the origin and thus little influenced by the sites (environmental conditions).
-The index of variability has almost no heritability.
-It is not surprising that this trait has very low heritability because it is estimated from model residuals after subtracting the typical density profile.
-Our study includes limitations for identifying families that may have low variability in wood density as part in tree improvement programs.
+Since parameters $b_1$ and $b_2$ are the parameters that describe the change in density over the years of tree growth, they are more likely to be influenced by environmental conditions.
+The index V has a heritability close to zero for each of the sites (Asselin 6.10^-6^ and Saint-Casimir 3.10^-6^) and the two sites combined (1.10^-7^).
+The low genetic control of the index V shows that in addition to the fact that the interannual density variability is controlled by environment [@becker1995; @huang2007; @Nicault2010; @zaehle2014], its amplitude is also controlled by environment.
+One hypothesis that can explain the low heritability of the V index is the young age of our trees.
+Indeed, the study of @Gaspar2008 showed an increase in the heritability of the components of density with age.
+As a result, we can hope for an increase in the heritability of the index V, which represents the amplitude of the interannual variation in wood density.
+
+Our study can allows identify trees with low variation from pith-to-bark but includes limitations for identifying tree that may have low interannual variation in wood density as part in tree genetic selection.
 Future studies should focus on the resilience and variability associated with one-time events such as short-term water stress.

### 2019-10-15 Article_AS.Rmd
Discussion

@@
 ---
-title: "On the eletrodynamic of moving bodies"
-author: "André D. Soro, Patrick Lenz, Jean Bousquet, Mariana Hassegawa, Jean-Romain Roussel, Alexis Achim"
+title: "Genetic influence on wood density variation in  white-spruce (*Picea glauca*)"
+author: André D. Soro, Patrick Lenz, Jean Bousquet, Mariana Hassegawa, Jean-Romain
+  Roussel, Alexis Achim
+editor_options:
+  chunk_output_type: inline
 output:
   pdf_document: default
+  html_document:
+    df_print: paged
   word_document: default
 bibliography: biblio.bibtex
-editor_options: 
-  chunk_output_type: inline
 ---
@@
 
+Wood density and rings width was measured from wood core taken from the sampled trees.
 Wood cores were extracted to a thickness of 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground).
@@
 A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations.
-We had hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks.
+Our data has a hierarchical structure, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks.
 Unlike some previous studies [@franceschini2010; @gardiner2011] that used mixed models to describe the radial density profile, we adjusted the curve of our equation to each core for a better description of the profiles.
@@
 We used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites.
+This comparison will allow determine the interaction between the environment and the family effect.
 The elements for which we calculated Spearman’s rho are: V for each of the tree families between two study sites and the three parameters to estimate from the model.
@@
 ## Quantitative genetic analyses
+Analyses of variance and covariance for various wood density and tree growth traits studied were performed to estimate genetic parameters.
 Variance components for genetic analysis were estimated using ASReml 3 [@Butler2018] based on linear mixed-effects model using as follows:
@@
 A summary of index V values can be found in figure 8.
-The maximum value is 1155.8 in Asselin, and 746 for the Saint-Casimir site, which represents a huge difference.
+The mean value is 417.5 in Asselin, and 339.8 for the Saint-Casimir site, which represents a huge difference.
 \
@@
 |Rank correlation  |Rho  |P-value |
-|:-----------------|----:|:-------|
+|:-----------------|----:|-------:|
 |Density residus   | 0.57|8.7^-09^|
@@
 
-The dataset showed a typical radial pattern of wood density already observed in spruce trees in other researchers [@Alteyrac2006; @GARDINER2011; @Xiang2014] with the density initially decreasing outwards from the pith before slowly increasing towards an asymptotic value.
+Our data showed a typical radial pattern of wood density already observed in spruce trees in other researchers [@Alteyrac2006; @GARDINER2011; @Xiang2014] with the density initially decreasing outwards from the pith before slowly increasing towards an asymptotic value.
 This pattern corresponds to the type II pith-to-bark profile described by @panshin1980.
 The high density of wood adjacent to the pith has been attributed to high earlywood density and low latewood proportion [@alteyrac2007; @koubaa2007].
-In addition, in the first years of a tree's life, it produces "flexure wood" that is high density [@telewski1989].
+Indeed, the high density of earlywood in the first years of growth is explained by the fact that in the first years of a tree's life, it produces "flexure wood" that is high density [@telewski1989].
 This type of wood is formed because the low rigidity of the young trees that allows them to withstand environmental constraints without breaking.
@@
 The modeling of individual profiles is suitable for the description of the wood density of our type II data.
-The prediction of the model on all of the data that can be seen in Figures 3 and 4 is an example of the modeling that is done for individual profiles.
+The choice of an equation containing two separate components allowed enough flexibility to represent the large variation in density profiles.
+The first term in the equation is capable of describing the type II density pattern previously described, whilst the second term was included when it was necessary to model a density increase close to the bark.
+The prediction of the model on all data that can be seen in Figures 3 and 4 is an example of the modeling that is done for individual profiles.
+The relatively poor fit of the full model in this study could be explained by the fact that, despite its classification as a type II species, wood density profiles in white spruce are, in reality, extremely variable.
 The choice of modeling the variation of the density of each of the trees made it possible to adapt the model to the variations between the trees.
@@
 Indeed, the study of @Gaspar2008 showed an increase in the heritability of the components of density with age.
-As a result, we can hope for an increase in the heritability of the index V, which represents the amplitude of the interannual variation in wood density.
+As a result, we can hope for an increase in the heritability of the index V for much older trees.
+\
 
@@
 As for the heritability of ring density and width, it is similar to what can be found in previous studies.
+This study showed that the amplitude of the high frequency signal is variable between trees but that this variation is not controlled genetically.
 The application of the method used in this study to older trees can be interesting in order to make a comparison and see if the heritability is the same and if this index can be used in this case to select families of trees for genetic improvement.

### 2019-11-04 Article_AS.Rmd
Resultats

@@
 
-#3 Results
+# 3 Results
 
-The model we have developed has made it possible to adjust wood density for each core.
+## Mean density per ring and prediction curve of the model
+
+When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings.
+It declined from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to 375 kg/m^3^ for the Asselin site.
+Then, this decrease was followed by a slow increase until the twelfth year.
+\
+
+![Pith-to-bark radial patern of the density measured from the cores taken at breast height by ring.
+A) Density for the site of Asselin and B) Density for the site of Saint-Casimir.
+The lines indicate densities for individual stands of samples.
+The bold lines indicate the density predicted by the adjusted model for all trees.](graph_table/Density.pdf)
+
+|Sites             |    N|$R^2$ Adjusted|       E|E   |E%  |
+|:-----------------|----:|-------------:|-------:|---:|---:|
+|Asselin           | 1035|0.49          |0.000008|58.6|0.15|
+|Saint-Casimir     | 1162|0.49          |0.00007 |48.1|0.13|
+
+The fit of the model we developed allowed us to take into account the density profile of each individual tree.
 A summary of the parameter values obtained of the model fitting process (Eq.
@@
 
-![Summary graphic of parameters in Asselin.](graph_table/parameter.asselin.png){width=335px; height=330px}
-
-![Summary graphic of parameters in Saint-Casimir.](graph_table/parameters.casimir.png){width=335px; height=330px}
-
-![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Asselin site.
-The bold lines shows the density predicted by Model.](graph_table/Density_asselin.png){width=330px; height=330px}
-
-![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model.](graph_table/Density_casimir.png){width=330px; height=330px}
-
-When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings.
-It declined from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to 375 kg/m^3^ for the Asselin site.
-Then, this decrease was followed by a slow increase until the twelfth year.
-\
+![Comparison of each parameter ($b_0$, $b_1$ and $b_2$) of equation 1 between the two study sites.
+A) Parameter $b_0$ denotes the intercept when t approaches a theoretical value of 0.
+B) Parameter $b_1$ denotes the scale parameter of the early decrease in ring density.
+C) Parameter $b_2$ denotes the rise in ring density at the end of the juvenile wood and at the beginning of the mature wood formation.
+The error bars represent standard deviation.](graph_table/summary_parameter)
 
@@
 
-<!--
-```{r aa, echo=FALSE}
-data <- readr::read_csv("data/Tableau_resume_indice d'uniformite.csv")
-knitr::kable(data)
-```
--->
-
-![Summary graphic of index V per site.](graph_table/indexV.png){width=335px; height=330px}
+![The index of the amplitude of variation of the interannual density (index V), determined from equation 2.
+A) Average amplitude of the variation of the interannual density at the Asselin site.
+B) Average amplitude of variation in interannual density at Saint-Casimir site.
+The bars represent the standard deviation.](graph_table/IndexV)
 
@@
 
-|Rank correlation  |Rho  |P-value |
+|Variables         |    R|P-value |
 |:-----------------|----:|-------:|
@@
 |$b_2$             | 0.60|2.2^-16^|
-Table: Sperman’s Rang correlation
+Table: Spearman rank correlation coefficients and their P-values between the two study sites (Asselin and Saint-Casimir) of the density, the index of the amplitude of interannual variation of the density and parameters of the equation 1.
 
@@
 
-![Abstract of the heritability ($hi^2$) of the density, the ring width, parameters $b_0$, $b_1$, $b_2$, index V and standard error per site.](graph_table/heritability.png){width=335px; height=330px}
+![Individual inheritance of: Ring density, ring width, index of the interannual variation of density (Index_V), parameters ($b_0$, $b_1$ and $b_2$) of equation 1, earlywood density per ring (ew_density), earlywood percentage per ring (ew_percentage), earlywood width (ew_width), latewood density (lw_density), ring index interannual variation in latewood density (lw_index_V), latewood percent per ring (lw_percentage) and latewood width per ring (lw_width) for each of our two study sites.
+A) The traits of the entire ring per site, B) earlywood wood traits by ring and site, C) latewood traits by ring and site.
+The bars represent the standard error.](graph_table/heritability.pdf)
 

### 2019-12-05 Article_AS.Rmd
Résultats

--- 2019-10-15 Article_AS.Rmd
@@
   pdf_document: default
+  word_document: default
   html_document:
     df_print: paged
-  word_document: default
 bibliography: biblio.bibtex
@@
 A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations.
+Latewood density in white spruce is more sensitive to annual variations in climate than the width of the ring [@darrigo1992].
+So we applied our model to density and percentage of latewood in order to better evaluate the amplitude of interannual variations in density.\
 Our data has a hierarchical structure, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks.
@@
 Where V is the index, $D_n$ is the wood density residues of year $n$ and  $D_{n - 1}$ is the wood density residues of year $n-1$. 
+V was determined for wood density over ring the width, density and percentage of latewood.
+Evaluation of this index for both ring width and latewood will compare the effects of the annual climate that influence the latewood and those associated with the spring conditions that influence ring density [@darrigo1992].
 
@@
 where $Y_{ijkl}$  is the observation on the $ijklth$ tree, $\mu$ is the overall mean, $s_i$ is the fixed site effect, $b_{(i)j}$ is the fixed effect of the $j^{th}$ bloc in site $i$, $f_k$ is the random effect of the $k^{th}$ family, $bf_{(i)j}$ is the random interaction of family and bloc, and $e_{ijkl}$ is the residual effect.
+The genetic control of these wood density components, weighted in each ring by their respective sectional area, was evaluated by estimating individual-tree heritability (h^2^).  
 In a full-sib design with random mating, members of the same family are presumed to share half of their genes [@white2007].
@@
 
-Where $hi^2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_t$ is total variance.
+Where $hi^2$ is individual heritability, $\sigma^2_f$ is family variance, $\sigma^2_t$ is total variance. \
 
-#3 Results
+# 3 Results
 
-The model we have developed has made it possible to adjust wood density for each core.
-A summary of the parameter values obtained of the model fitting process (Eq.
-1) can be found in figure 4&5.
-We observe that the average value of the model parameters differs between Asselin and Saint-Casimir.
-For the parameters $b_0$ and $b_1$, the average is higher in Saint-Casimir and for parameter $b_2$ it is higher in Asselin.
-Although there are small differences in the parameters between the two sites, these differences are very small.
-The fit of the model is therefore similar at both sites.
-
-![Summary graphic of parameters in Asselin.](graph_table/parameter.asselin.png){width=335px; height=330px}
-
-![Summary graphic of parameters in Saint-Casimir.](graph_table/parameters.casimir.png){width=335px; height=330px}
-
-![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Asselin site.
-The bold lines shows the density predicted by Model.](graph_table/Density_asselin.png){width=330px; height=330px}
-
-![Pith-to-bark radial pattern of breast height density by ring.
-Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model.](graph_table/Density_casimir.png){width=330px; height=330px}
-
-When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings.
-It declined from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to 375 kg/m^3^ for the Asselin site.
-Then, this decrease was followed by a slow increase until the twelfth year.
+## Mean density per ring and model prediction
+Wood density variation from the pith-to-bark is similar in the two study sites between the first and the fifth year (Fig.
+4).
+The pattern observed at the two sites corresponds to type II pith-to-bark profile described in introduction.
+When considering all the cores of trees from the pith-to-bark, ring density decreased rapidly from 430 kg/m^3^ to 370 kg/m^3^ for Saint-Casimir and from 455 kg/m^3^ to 375 kg/m^3^ for Asselin site in the first 5 growth rings.
+From the sixth year, we observe a stabilization of the density at 375 kg/m^3^ of the site of Asselin (Fig.
+4A) unlike the site of Saint-Casimir (Fig.
+4B) where we observe a increase  to reach a plateau at the eigth year (380 kg/m^3^).
 \
 
-Model predictions are similar to data trends across sites (figs 6 and 7).
-A decrease in density is observed in the first years of growth and then an increase in density from a certain cambial age.
-Whether it is the data or the predictions of the model, we observe a difference between the Asselin (fig 6) and Saint-Casimir (fig 7) sites.
-The increase in density after the decrease of the first years of growth is greater in Saint-Casimir than in Asselin.
-However, the average density is very similar between the two sites.
-In spite of the interannual variations, the trees have the same general tendency and this one seems well described by the model which adapts to each of the sites.\
-
-<!--
-```{r aa, echo=FALSE}
-data <- readr::read_csv("data/Tableau_resume_indice d'uniformite.csv")
-knitr::kable(data)
-```
--->
-
-![Summary graphic of index V per site.](graph_table/indexV.png){width=335px; height=330px}
-
-When comparing the index V between study sites, there is greater variability in Asselin than in Saint-Casimir.
-A summary of index V values can be found in figure 8.
-The mean value is 417.5 in Asselin, and 339.8 for the Saint-Casimir site, which represents a huge difference.
+Like ring density, the latewood density has a type II pith-to-bark profile for two sites (Fig.
+4C: Asselin, Fig.
+4D: Saint-Casimir).
+Latewood density decreased rapidly from 500 kg/m^3^ to 480 kg/m^3^ for Saint-Casimir and from 510 kg/m^3^ to 450 kg/m^3^ for Asselin site in the first 5 growth rings.
+From the sixth year, we observe a stabilization of the density at 450 kg/m^3^ of the site of Asselin (Fig.
+4C) unlike the site of Saint-Casimir (Fig.
+4D) where we observe a increase  to reach a plateau at the eigth year (4900 kg/m^3^).
 \
 
-|Rank correlation  |Rho  |P-value |
-|:-----------------|----:|-------:|
-|Density residus   | 0.57|8.7^-09^|
-|Index V           | 0.08|0.43    |
-|$b_0$             | 0.70|2.2^-16^|
-|$b_1$             | 0.80|2.2^-16^|
-|$b_2$             | 0.60|2.2^-16^|
-Table: Sperman’s Rang correlation
+As for the percentage of latewood, the trend has a type II pith-to-bark profile for two sites  (Fig.
+4E: Asselin, Fig.
+4F: Saint-Casimir).
+In Asselin, the percentage of latewood decreases between 28% in the first year and 15% in the fourth year, then increases to 17% before decreasing again to 12.5%.
+In Saint-Casimir, latewood decreases between 25% in the first year and 12.5% in the fourth year, then increases to 15% before decreasing again to 13%.\
 
-The results of the rank correlation coefficient are summarized in Table 2.
-These results show an absence of correlation for the index V (p-value = 0.43) and a strong correlation for all the parameters of the model (p-value = 2.2 10^-16^).
-The value of Rho is very similar between the three parameters.
+![Pith-to-bark radial patern of the density measured from the cores taken at breast height by ring (A and B).
+A) Density for the site of Asselin and B) Density for the site of Saint-Casimir.
+C and D represent Pith-to-bark radial patern of latewood density (C : Asselin and D: Saint-Casimir).
+E and F represent pith-to-bark radial patern of latewood percentage (E: Asselin and F: Saint-Casimir).
+The bold lines indicate the density predicted by the adjusted model for all trees.](graph_table/density2.pdf){height=80%}
+
+|Sites             |N    |$R^2$ Adjusted|       E|  E    |  E%|
+|:-----------------|----:|-------------:|-------:|------:|---:|
+|Asselin           | 1035|0.49          |0.000008|58.6   |0.15|
+|Saint-Casimir     | 1162|0.49          |0.00007 |48.1   |0.13|
+
+Table: Fit indices and error statistics for the fitted model. Mean error (E), mean absolute error (|E|), and mean absolute percentage error (E%) were calculated from the fixed part of each model.
+
+The fit of the model we developed allowed us to take into account the density profile of each individual tree.
+A summary of the parameter values obtained of the model fitting process (Eq.
+1) can be found in table 3.
+The parameters of the model are similar between the two sites, the standard deviations of the mean of these intersect (Table 3).
+Non-significance of the difference that exists between the parameters is observed at the level of those of the ring density , those of the latewood density and those of the percentage of latewood.
+The percent change in density explained by cambial age is 49% in the model for each of our two sites (Equation 1) (Table 2).
+The mean error, the mean absolute error, and the mean percent error were 8^-6^ $kg.m^-3$, 58.6 $kg.m^-3$ and 15%, respectively, for the Asselin site and 7^-5^ $kg.m^-3$, 48.1 $kg.m^-3$ and 13% for the Saint-Casimir site (Table 2).
+There were no obvious trends in plots of the normalized residuals versus fitted and explanatory variables for Equation 1 (results not shown).
+The fit of the model is therefore similar at both sites (49%).
 \
 
-![Abstract of the heritability ($hi^2$) of the density, the ring width, parameters $b_0$, $b_1$, $b_2$, index V and standard error per site.](graph_table/heritability.png){width=335px; height=330px}
+|Parameters          |Asselin|Saint-Casimir|
+|:------------------:|------:|------:|
+| $b_0$              |302.78 $\pm$ 55.90|304.40 $\pm$ 41.90|
+| $b_1$              |0.97 $\pm$  0.16 |1.09 $\pm$ 0.15 |
+| $b_2$              |17.02 $\pm$ 2.20 |14.87 $\pm$  2.00 |
+| $b_0$_lw_density   |568.02 $\pm$ 70.30|443.30 $\pm$ 53.00|
+| $b_1$_lw_density   |-0.04 $\pm$ 0.19 |0.10 $\pm$ 0.10 |
+| $b_2$_lw_density   |-299.10 $\pm$ 30.31|-268.40 $\pm$ 200.0|
+| $b_0$_lw_percentage|68.82 $\pm$ 30.3 |85.80 $\pm$ 30.80|
+| $b_1$_lw_percentage|-104.60 $\pm$ 33.04|-133.40 $\pm$ 21.80|
+| $b_2$_lw_percentage|79.69 $\pm$ 20.77|18.30 $\pm$ 2.46 |
+Table: Comparison of each parameter average ($b_0$, $b_1$ and $b_2$) of equation 1 between the two study sites.
+Parameter $b_0$ denotes the intercept when t approaches a theoretical value of 0.
+Parameter $b_1$ denotes the scale parameter of the early decrease in ring density.
+Parameter $b_2$ denotes the rise in ring density at the end of the juvenile wood and at the beginning of the mature wood formation.
+$b_0$_lw_density is $b_0$ of latewood density, $b_1$_lw_density is $b_1$ of latewood density, $b_2$_lw_density is $b_2$ of latewood density, $b_0$_lw_percentage is $b_0$ of latewood percentage, $b_1$_lw_percentage is $b_1$ of latewood percentage and $b_2$_lw_percentage is $b_0$ of latewood percentage.
 
-Heritability is higher in each site separately than that of the combined sites for all traits (fig 9).
-Wood density is the only character with strong heritability for each of the sites as well as for the combined sites.
-The annual ring width and the parameter $b_0$ have low heritability.
-When parameters $b_1$ and $b_2$, their heritability is very low.
-The index V has almost zero heritability regardless of the study site.\
+## Interannual variation in density and index V
+Interannual variation of the density can be observed in Figure 5.
+It has an amplitude that varies from one tree to another.
+This amplitude was measured from our index V.
+The average value of the index V (Asselin: 417.5 $kg.m^-3$, Saint-Casimir: 339.8 $kg.m^-3$) is represented in figure 5A.
+This result shows that the index V is not significantly (standard deviation) different between the two study sites.
+Trees therefore behave similarly between sites with regard to the V index.
+Thus this index allows a classification of the trees in order from the amplitude of interannual variation of density of the wood.
+The trees with the lowest V index are those with low interannual variation.
+The average value of the index V (Asselin: 293.7 $kg.m^-3$, Saint-Casimir: 292.1 $kg.m^-3$) determined from latewood is shown in Figure 5B.
+The average value of the index V (Asselin: 53.5 $kg.m^-3$, Saint-Casimir: 41.9 $kg.m^-3$) determined from latewood percentage is shown in Figure 5C.
+\
+
+![Index avarage of the amplitude of interannual variation of density (V), latewood density ($V_lw$) and latewood density percentage ($V_lw.percentage$) determined from equation 2.
+A) Average amplitude of the variation of the interannual density ring whole.
+B) Average amplitude of the variation of the interannual latewood density.
+C) Average amplitude of the variation of the interannual latewood density percentage.
+Standard errors of genetic correlations are given by error bars.](graph_table/index_V.pdf){width=85%, height=90%}
+
+\clearpage
+
+## Family-tree interaction and individual heritability.
+The results of the rank correlation coefficient are summarized in figure 6.
+These results show an absence of correlation for the index V (p > 0.05), a significant correlation for ring density (p < 10^-5^), a significant correlation for latewood density (p < 10^-5^), a significant correlation for latewood percentage (p < 10^-5^), a significant correlation for latewood index V (p < 10^-5^), a significant correlation for latewood percentage index V (p < 10^-5^), a significant correlation for all the parameters of the model (p < 10^-5^).
+The value of R is very similar between the parameters.
+There is a significant correlation for traits related variation in density from pith-to-bark and for latewood index V.
+On the other hand, there is no correlation when considering the interannual variation of the density.
+
+![Spearman rank correlation coefficients and their significativity between the two study sites (Asselin and Saint-Casimir) of the density, index of the amplitude of interannual variation of the density (V), index of the amplitude of interannual variation of latewood density (Vlwd), $b_0$ of latewood density ($b_0$.lw.d), $b_1$ of latewood density ($b_1$.lw.d), $b_2$ of latewood density ($b_2$.lw.d), $b_0$ of latewood percentage ($b_0$.lw.p), $b_1$ of latewood percentage ($b_1$.lw.p), $b_2$ of latewood percentage ($b_2$.lw.p) and parameters of the equation 1.
+Significance levels are indicated by ***: P<0.0001; NS:no significant.](graph_table/Spearman_rank_correlation.pdf)
+
+Individual inheritance of: Ring density, ring width , index of the interannual variation of density (V), parameters ($b_0$, $b_1$ and $b_2$) of equation 1, latewood density (lw.d), index interannual variation in latewood density (V.lw), $b_0$ of latewood density ($b_0$.lw), $b_1$ of latewood density ($b_1$.lw), $b_2$ of latewood density ($b_2$.lw), ring index interannual variation in latewood percentage (V.lp), latewood percent per ring (lp), $b_0$ of latewood percentage ($b_0$.lp), $b_1$ of latewood percentage ($b_1$.lp) and $b_2$ of latewood percentage ($b_2$.lp) for each of our two study sites is shown in Figure 7.
+This graph presents estimates of individual heritability for the individual tree level.
+Analysis of heritability showed no significant differences among sites for all traits.
+At first, when we consider the ring in its entirety, heritability varies between 6.10^-6^ and 0.58 (Asselin) and between 3.10^-6^ and 0.71 (Saint-Casimir) depending on the traits studied.
+Individual heritability is not significantly different from one to the other.
+On the other hand, it is significantly different between the traits that are related to the variation of the density of pith-to-bark and that of the amplitude of interannual variation.
+except index V of latewood percentage, all the parameters have a lower heritability for latewood percentage than for the density of latewood or the entire ring.
+\
+
+![Individual inheritance of: Ring density, index of the interannual variation of density (V), parameters ($b_0$, $b_1$ and $b_2$) of equation 1, latewood density (lw.d), index interannual variation in latewood density (V.lw), $b_0$ of latewood density ($b_0$.lw), $b_1$ of latewood density ($b_1$.lw), $b_2$ of latewood density ($b_2$.lw), ring index interannual variation in latewood percentage (V.lp),latewood percent per ring (lp), $b_0$ of latewood percentage ($b_0$.lp), $b_1$ of latewood percentage ($b_1$.lp) and $b_2$ of latewood percentage ($b_2$.lp) for each of our two study sites.
+Standard errors of genetic correlations are given by error bars.](graph_table/heritability2.pdf)
 
